# ENCE - Parallel vs Non-Parallel R Updates 
Checking that R's parallelism (mclapply) for station updates reproduces the same result as the serial version (with synchronous updates), and getting a first look at the speedup 

Imputing daily Met Eireann dataset, disregarding missing data and imputing on masked data to test RMSE

The original ENCE algorithm imputes each column of the data matrix $Z$ column-by-column within an outer cycle, where each station's regression model is fit using the most recently imputed values of all other stations.

The alternative algorithm considered here updates all stations within a cycle against a frozen snapshot of $Z$ taken at the start of that cycle, so the update order within a cycle no longer matters. The motivation for this change is to allow for parallelism.

Further details of ENCE can be found in the study at <https://doi.org/10.1002/joc.8513> The data that support the findings of this study are available upon reasonable request to the author(s) and with the permission of Met Éireann, Ireland's meteorological office.

## Setup

Load the Met Eireann daily data (not available in this Github repo)

```{r setup, message=FALSE}
source("idw.R")
source("ENCE_parallel.R")      # synchronous (parallel) updates
source("ENCE_synchronous.R")   # synchronous (non-parallel) updates

# Load Met Eireann daily data and station geodata file
rain     <- read.csv("../met_eireann_daily_data/daily_rainfall_2016.csv")
stations <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

head(rain)
str(rain)

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)#
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Adds a completeness column = observed days / 366.
obs_per_station$completeness <- obs_per_station$n_obs / 366

# Apply the % cutoff
cutoff <- 0.50
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Then mask a fraction of the currently-observed cells
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

truth   <- rain$y                     # ground truth at masked cells
df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```

## Imputation

Run both variants on the daily data:

```{r run-parallel_and_sychronous, message=FALSE, results='hide'}
source("ENCE_parallel.R")
t0 <- Sys.time()
imp_par <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = truth, masked_idx = masked_idx)
rmse_par     <- attr(imp_par, "rmse_history")
ext_rmse_par <- attr(imp_par, "ext_rmse_history")
t_par <- Sys.time() - t0
cat(sprintf("Parallel: %s\n", format(t_par)))

t0 <- Sys.time()
source("ENCE_synchronous.R")
imp_sync <- ENCE(df_miss, response = "y", max_cycles = 30,
                 init_method = "idw",
                 x_name = "east", y_name = "north", t_name = "t",
                 C = 15000,
                 transformation = sqrt,
                 reverse_transformation = function(x) x^2,
                 truth = truth, masked_idx = masked_idx)
rmse_sync <- attr(imp_sync, "rmse_history")
ext_rmse_sync <- attr(imp_sync, "ext_rmse_history")
t_ser <- Sys.time() - t0
cat(sprintf("Non-Parallel: %s\n", format(t_ser)))
```

## Check
```{r reproducibility-and-speedup}
# Bit-identical check
ident_ok <- identical(imp_par$y, imp_sync$y)
max_diff <- max(abs(imp_par$y - imp_sync$y), na.rm = TRUE)

cat(sprintf("identical(parallel, non-parallel): %s\n", ident_ok))
cat(sprintf("Max abs difference:                %g\n", max_diff))

# Speedup
cat(sprintf("Speedup: %.2fx on %d cores\n",
            as.numeric(t_ser, units = "secs") / as.numeric(t_par, units = "secs"),
            parallel::detectCores() - 1))
```


Plot 1: Internal Convergence

```{r plot-convergence, fig.width=8, fig.height=5}
xlim_range <- c(1, max(length(rmse_par), length(rmse_sync)))

plot(rmse_par, type = "b", col = "blue", pch = 16,
     xlim = xlim_range,
     ylim = range(c(rmse_par, rmse_sync)),
     xlab = "Cycle", ylab = "Internal RMSE (between cycles)",
     main = "Internal convergence: Parallel vs Non-Parallel")
lines(rmse_sync, type = "b", col = "red", pch = 17)
legend("topright", legend = c("Parallel", "Non-Parallel"),
       col = c("blue", "red"), pch = c(16, 17), lty = 1)
```

Plot 2: RMSE Validation
```{r plot-RMSE, fig.width=8, fig.height=5}
xlim_range <- c(1, max(length(ext_rmse_par), length(ext_rmse_par)))

plot(ext_rmse_par, type = "b", col = "blue", pch = 16,
     xlim = xlim_range,
     ylim = range(c(ext_rmse_par, ext_rmse_sync)),
     xlab = "Cycle", ylab = "Validation RMSE (masked cells)",
     main = "Validation RMSE: Parallel vs Non-parallel")
lines(ext_rmse_sync, type = "b", col = "red", pch = 17)
legend("topright", legend = c("Parallel", "Synchronous"),
       col = c("blue", "red"), pch = c(16, 17), lty = 1)
```



## Data summary

```{r data-summary}
cat(sprintf("Stations: %d\n", length(unique(rain$stno))))
cat(sprintf("Rows:     %d\n", nrow(rain)))
cat(sprintf("NAs:      %d (%.1f%%)\n", sum(is.na(rain$y)), 100 * mean(is.na(rain$y))))

na_per_station <- aggregate(is.na(rain$y), list(stno = rain$stno), sum)
summary(na_per_station$x)

cat(sprintf("Stations: %d\n", length(unique(rain$stno))))
cat(sprintf("Rows:     %d\n", nrow(rain)))
cat(sprintf("Real NAs:    %d (%.1f%%)\n", sum(is.na(rain$y)), 100 * mean(is.na(rain$y))))
cat(sprintf("Masked:      %d (%.1f%% of observed)\n", length(masked_idx), 100 * length(masked_idx) / sum(!is.na(rain$y))))
cat(sprintf("ENCE sees:   %d NAs total (%.1f%%)\n", sum(is.na(df_miss$y)), 100 * mean(is.na(df_miss$y))))
```