# Filename: imputation_run.Rmd
# Author: Marc Lane

# Final Imputation Run - Computational Time Comparison

Run this .Rmd to reproduce the results from section 4.3 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/ENCE_C.R")
source("scripts/ENCE_elastic_net_C_implementation/elastic_net_R_glue.R")
source("scripts/MICE_C.R")
source("scripts/MICE_elastic_net_C_implementation/elastic_net_R_glue.R")
source("scripts/PPCA.R")
source("scripts/MissForest.R")
source("scripts/MLR.R")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_3_Imputation_Run_%s.rds", tag)
log_file     <- sprintf("section_4_3_Imputation_Run_%s.txt", tag)
```

# Imputation

Run every method once per rep on the full 10 years of data, exactly as-is (no masking, no held-out truth). 
Repeat 3 times to measure timing variance

```{r run, message=FALSE, results='hide'}
n_reps <- 3

# Per-rep timing (parallel to the per-fold time_vec in the other files, now indexed by rep)
ence_time_vec <- numeric(n_reps)
mice_time_vec <- numeric(n_reps)
idw_time_vec  <- numeric(n_reps)
ppca_time_vec <- numeric(n_reps)
mf_time_vec   <- numeric(n_reps)
mlr_time_vec  <- numeric(n_reps)

# Number of genuinely-missing cells actually filled by each method per rep
ence_n_imputed <- numeric(n_reps)
mice_n_imputed <- numeric(n_reps)
idw_n_imputed  <- numeric(n_reps)
ppca_n_imputed <- numeric(n_reps)
mf_n_imputed   <- numeric(n_reps)
mlr_n_imputed  <- numeric(n_reps)

missing_idx <- which(is.na(rain$y))

for (r in seq_len(n_reps)) {
  cat("Rep", r, "of", n_reps, "\n")
  set.seed(3000 + r)

  df_miss <- rain

  # Run ence
  ence_time_vec[r] <- system.time({
    ence_imp <- ENCE_C(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                lambda_options = lambda_options,
                alpha_options  = alpha_options
                )
  })["elapsed"]
  ence_n_imputed[r] <- sum(!is.na(ence_imp$y[missing_idx]))


  # Run mice
  mice_time_vec[r] <- system.time({
    mice_imp <- MICE_DURR_C(df_miss, response = "y", m = 4,
                       max_impute_cycles = 5, hyp_cycles = 2,
                       init_method = "idw", max_EM_cycles = 8,
                       nclusters = 1, nthreads = 16, outfile = NULL,
                       lambda_options = lambda_options,
                       alpha_options  = alpha_options,
                       transformation = sqrt,
                       reverse_transformation = function(x) x^2)
  })["elapsed"]
  mice_n_imputed[r] <- sum(!is.na(mice_imp$y[missing_idx]))


  # Run IDW
  obs    <- df_miss[!is.na(df_miss$y), ]
  target <- df_miss[is.na(df_miss$y), ]
  idw_time_vec[r] <- system.time({
    idw_imp <- IDW_ST(obs, target, response = "y",
                  x_name = "east", y_name = "north", t_name = "t",
                  C = 15000,
                  transformation = sqrt,
                  reverse_transformation = function(x) x^2)
  })["elapsed"]
  idw_n_imputed[r] <- sum(!is.na(idw_imp$y))


  # Run PPCA
  ppca_time_vec[r] <- system.time({
    ppca_imp <- PPCA(df_miss, nPcs = 20, transformation = sqrt,
                   reverse_transformation = function(x) x^2)
  })["elapsed"]
  ppca_n_imputed[r] <- sum(!is.na(ppca_imp$y[missing_idx]))


  # Run MissForest
  mf_time_vec[r] <- system.time({
    mf_imp <- MissForest(df_miss, ntree = 100, mtry = 100,
                     transformation = sqrt, reverse_transformation = function(x) x^2)
  })["elapsed"]
  mf_n_imputed[r] <- sum(!is.na(mf_imp$y[missing_idx]))


  # Run MLR
  obs    <- df_miss[!is.na(df_miss$y), ]
  target <- df_miss[is.na(df_miss$y), ]
  mlr_time_vec[r] <- system.time({
    mlr_imp <- impute_MLR(obs, target, nn = 5, response = "y",
                    transformation = sqrt, reverse_transformation = function(x) x^2)
  })["elapsed"]
  mlr_n_imputed[r] <- sum(!is.na(mlr_imp$y[missing_idx]))
}

# Totals across the 3 reps
ence_time <- sum(ence_time_vec)
mice_time <- sum(mice_time_vec)
idw_time  <- sum(idw_time_vec)
ppca_time <- sum(ppca_time_vec)
mf_time   <- sum(mf_time_vec)
mlr_time  <- sum(mlr_time_vec)
```

## Helper Functions
```{r helper functions, message=FALSE}
report_timing <- function(method_name, elapsed_vec, n_missing, n_imputed_vec) {
  cat(method_name, "\n", sep = "")
  cat("Mean time:", round(mean(elapsed_vec), 2), "seconds  (sd:", round(sd(elapsed_vec), 2),
      "over", length(elapsed_vec), "reps)\n")
  cat("Per-rep times:", paste(round(elapsed_vec, 2), collapse = ", "), "\n")
  cat("Coverage:", round(mean(n_imputed_vec), 1), "of", n_missing,
      sprintf("(%.1f%%)\n\n", 100 * mean(n_imputed_vec) / n_missing))
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "Final Imputation Run - computational time",
    cutoff      = cutoff,
    n_reps      = n_reps,
    years       = years,
    n_stations  = length(keep),
    n_missing   = length(missing_idx),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all 3 reps
  ence_time = ence_time,
  mice_time = mice_time,
  idw_time = idw_time,
  ppca_time = ppca_time,
  mf_time = mf_time,
  mlr_time = mlr_time,
  ence_time_vec = ence_time_vec,
  mice_time_vec = mice_time_vec,
  idw_time_vec = idw_time_vec,
  ppca_time_vec = ppca_time_vec,
  mf_time_vec = mf_time_vec,
  mlr_time_vec = mlr_time_vec,

  # Coverage bookkeeping (fraction of genuinely-missing cells actually filled per rep)
  ence_n_imputed = ence_n_imputed,
  mice_n_imputed = mice_n_imputed,
  idw_n_imputed = idw_n_imputed,
  ppca_n_imputed = ppca_n_imputed,
  mf_n_imputed = mf_n_imputed,
  mlr_n_imputed = mlr_n_imputed
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_timing("ENCE", ence_time_vec, length(missing_idx), ence_n_imputed)
  report_timing("MICE", mice_time_vec, length(missing_idx), mice_n_imputed)
  report_timing("IDW", idw_time_vec, length(missing_idx), idw_n_imputed)
  report_timing("PPCA", ppca_time_vec, length(missing_idx), ppca_n_imputed)
  report_timing("MissForest", mf_time_vec, length(missing_idx), mf_n_imputed)
  report_timing("MLR", mlr_time_vec, length(missing_idx), mlr_n_imputed)
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```
