# Filename: imputation_run_sklearn.Rmd
# Author: Marc Lane

# Final Imputation Run - Computational Time Comparison

Run this .Rmd to reproduce the results from section 4.3 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

# Use reticulate to call sklearn from R
python_path <- Sys.getenv("SKLEARN_PYTHON", unset = "/path/to/python3")
reticulate::use_python(python_path, required = TRUE)
reticulate::source_python("scripts/iterative_imputer.py")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")


# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_3_Imputation_Run_sklearn_%s.rds", tag)
log_file     <- sprintf("section_4_3_Imputation_Run_sklearn_%s.txt", tag)
```

# Imputation

Run every method once per rep on the full 10 years of data, exactly as-is (no masking, no held-out truth). 
Repeat 3 times to measure timing variance

```{r run, message=FALSE, results='hide'}
n_reps <- 3

sklearn_time_vec <- numeric(n_reps)
sklearn_n_imputed <- numeric(n_reps)

missing_idx <- which(is.na(rain$y))

stno_levels  <- sort(unique(rain$stno))
t_levels     <- sort(unique(rain$t))
row_idx_full <- match(rain$t, t_levels)
col_idx_full <- match(rain$stno, stno_levels)

for (r in seq_len(n_reps)) {
  cat("Rep", r, "of", n_reps, "\n")
  set.seed(3000 + r)

  df_miss <- rain

  X_matrix <- matrix(NA_real_, nrow = length(t_levels), ncol = length(stno_levels))
  row_idx  <- match(df_miss$t, t_levels)
  col_idx  <- match(df_miss$stno, stno_levels)
  X_matrix[cbind(row_idx, col_idx)] <- df_miss$y

  sklearn_time_vec[r] <- system.time({
    X_imputed <- run_iterative_imputer(X_matrix, n_nearest_features = 50L,
                                        max_iter = 10L, random_state = 916L)
  })["elapsed"]

  pred_all <- X_imputed[cbind(row_idx_full, col_idx_full)]
  sklearn_n_imputed[r] <- sum(!is.na(pred_all[missing_idx]))
}

sklearn_time <- sum(sklearn_time_vec)
```

## Helper Functions
```{r helper functions, message=FALSE}
report_timing <- function(method_name, elapsed_vec, n_missing, n_imputed_vec) {
  cat(method_name, "\n", sep = "")
  cat("Mean time:", round(mean(elapsed_vec), 2), "seconds  (sd:", round(sd(elapsed_vec), 2),
      "over", length(elapsed_vec), "reps)\n")
  cat("Per-rep times:", paste(round(elapsed_vec, 2), collapse = ", "), "\n")
  cat("Coverage:", round(mean(n_imputed_vec), 1), "of", n_missing,
      sprintf("(%.1f%%)\n\n", 100 * mean(n_imputed_vec) / n_missing))
}
```

# Save Results
```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "sklearn ONLY, Final Imputation Run, computational time",
    cutoff      = cutoff,
    n_reps      = n_reps,
    years       = years,
    n_stations  = length(keep),
    n_missing   = length(missing_idx),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all 3 reps
  sklearn_time = sklearn_time,
  sklearn_time_vec = sklearn_time_vec,

  sklearn_n_imputed = sklearn_n_imputed
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_timing("sklearn", sklearn_time_vec, length(missing_idx), sklearn_n_imputed)
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```
