# ENCE - C Implementation Test

In the directory `elastic_net_C_implementation`, I implemented my own version of `glmnet` (elastic net regressor) written in C (this will allow me to use OpenMP later). 

This file investigates if the C implementation of the coordinate descent recreates the same result as glmnet
## Setup

Load the Met Eireann daily data (not available in this Github repo)

```{r setup, message=FALSE}
source("idw.R")
source("elastic_net_C_implementation/elastic_net_R_glue.R")

# Load Met Eireann daily data and station geodata file
rain     <- read.csv("../met_eireann_daily_data/daily_rainfall_2016.csv")
stations <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)#
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Adds a completeness column = observed days / 366.
obs_per_station$completeness <- obs_per_station$n_obs / 366

# Apply the % cutoff
cutoff <- 0.50
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Then mask a fraction of the currently-observed cells
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```

## Imputation

Run the original `glmnet`-based ENCE first, then the C-version on the same masked data.

```{r run-glmnet, message=FALSE, results='hide'}
source("ENCE_glmnet.R")
time_glmnet <- system.time({
  imp_glmnet <- ENCE(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = masked_idx)
})
ext_rmse_glmnet <- attr(imp_glmnet, "ext_rmse_history")
```

```{r run-c, message=FALSE, results='hide'}
source("ENCE_myelasticnet.R")
time_c <- system.time({
  imp_c <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx)
})
ext_rmse_c <- attr(imp_c, "ext_rmse_history")
```

## Results

### Timing

```{r timing}
cat("glmnet version:   ", round(time_glmnet["elapsed"], 2), "seconds\n")
cat("C version:        ", round(time_c["elapsed"], 2),       "seconds\n")
```

### Accuracy against truth

Final external RMSE (predicted vs. true rainfall at the masked locations):

```{r final-rmse}
cat("glmnet final ext_rmse: ", round(tail(ext_rmse_glmnet, 1), 4), "\n")
cat("C final ext_rmse: ", round(tail(ext_rmse_c, 1), 4),       "\n")
```

Per-cycle external RMSE trajectory:

```{r rmse-plot, fig.width=8, fig.height=5}
xlim_range <- c(1, max(length(ext_rmse_glmnet), length(ext_rmse_c)))
plot(ext_rmse_glmnet, type = "b", col = "blue", pch = 16,
     xlim = xlim_range,
     ylim = range(c(ext_rmse_glmnet, ext_rmse_c)),
     xlab = "Cycle", ylab = "Validation RMSE",
     main = "Validation RMSE convergence: glmnet vs C")
lines(ext_rmse_c, type = "b", col = "red", pch = 17)
legend("topright", legend = c("glmnet", "C"),
       col = c("blue", "red"), pch = c(16, 17), lty = 1)
```

### Agreement between the two implementations

```{r imputation-agreement}
pred_glmnet <- as.numeric(unlist(imp_glmnet[masked_idx, "y"]))
pred_c <- as.numeric(unlist(imp_c[masked_idx, "y"]))
imp_diff_rmse <- sqrt(mean((pred_glmnet - pred_c)^2))
cat("RMSE between glmnet and C imputations at masked rows:", round(imp_diff_rmse, 4), "\n")
```

### Calculating Bias and R^2
```{r R^2 and Bias}
truth_masked <- rain$y[masked_idx]
pred_glmnet <- as.numeric(unlist(imp_glmnet[masked_idx, "y"]))
pred_c <- as.numeric(unlist(imp_c[masked_idx, "y"]))

bias_glmnet <- mean(pred_glmnet - truth_masked)
bias_c <- mean(pred_c - truth_masked)

ss_tot <- sum((truth_masked - mean(truth_masked))^2)
r2_glmnet <- 1 - sum((truth_masked - pred_glmnet)^2) / ss_tot
r2_c <- 1 - sum((truth_masked - pred_c)^2)      / ss_tot

cat("glmnet bias: ", round(bias_glmnet, 4), "\n")
cat("C bias: ", round(bias_c, 4), "\n")
cat("glmnet R^2: ", round(r2_glmnet,   4), "\n")
cat("C R^2: ", round(r2_c, 4), "\n")
```