
# Filename: methods_test.Rmd
# Author: Marc Lane

# Methods Comparison Test

Run this .Rmd to reproduce the results from the first part of Section 4.3 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/ENCE_C.R")
source("scripts/ENCE_elastic_net_C_implementation/elastic_net_R_glue.R")
source("scripts/MICE_C.R")
source("scripts/MICE_elastic_net_C_implementation/elastic_net_R_glue.R")
source("scripts/PPCA.R")
source("scripts/MissForest.R")
source("scripts/MLR.R")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_3_Methods_Comparison_%s.rds", tag)
log_file     <- sprintf("section_4_3_Methods_Comparison_%s.txt", tag)
```

# Imputation
 
Split data into 10 folds, mask observed entries and impute for each fold separately
```{r run, message=FALSE, results='hide'}
# Split the observed cells into K-folds
set.seed(916)
K <- 10
# Indexes of observed cells
observed_idx <- which(!is.na(rain$y))
# Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
fold_id      <- sample(rep(1:K, length.out = length(observed_idx)))
# Groups observed_idx into its fold
folds        <- split(observed_idx, fold_id)

# Initialising variables
ence_pred <- rep(NA_real_, nrow(rain))
mice_pred <- rep(NA_real_, nrow(rain))
idw_pred  <- rep(NA_real_, nrow(rain))
ppca_pred <- rep(NA_real_, nrow(rain))
mf_pred  <- rep(NA_real_, nrow(rain))
mlr_pred  <- rep(NA_real_, nrow(rain))
ence_time <- 0
mice_time <- 0
idw_time <- 0
ppca_time <- 0
mf_time  <- 0
mlr_time  <- 0
ence_time_vec <- numeric(K)
mice_time_vec <- numeric(K)
idw_time_vec  <- numeric(K)
ppca_time_vec  <- numeric(K)
mf_time_vec  <- numeric(K)
mlr_time_vec  <- numeric(K)

# Loop through each fold
for (k in seq_len(K)) {
  cat("Fold", k, "of", K, "\n")
  
  test_idx <- folds[[k]]
  
  # Hide the masked data for this fold
  df_miss <- rain
  df_miss$y[test_idx] <- NA
  
  # Run ence
  ence_time_vec[k] <- system.time({
    ence_imp <- ENCE_C(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = test_idx,
                lambda_options = lambda_options,
                alpha_options  = alpha_options
                )
  })["elapsed"]
  ence_time <- ence_time + ence_time_vec[k]


  # Run mice
  mice_time_vec[k] <- system.time({
    mice_imp <- MICE_DURR_C(df_miss, response = "y", m = 4,
                       max_impute_cycles = 5, hyp_cycles = 2,
                       init_method = "idw", max_EM_cycles = 8,
                       nclusters = 1, nthreads = 16, outfile = NULL,
                       lambda_options = lambda_options,
                       alpha_options  = alpha_options,
                       transformation = sqrt,
                       reverse_transformation = function(x) x^2)
  })["elapsed"]
  mice_time <- mice_time + mice_time_vec[k]


  # Run IDW
  obs    <- df_miss[!is.na(df_miss$y), ]
  target <- rain[test_idx, ]
  idw_time_vec[k] <- system.time({
    idw_imp <- IDW_ST(obs, target, response = "y",
                  x_name = "east", y_name = "north", t_name = "t",
                  C = 15000,
                  transformation = sqrt,
                  reverse_transformation = function(x) x^2)
  })["elapsed"]
  idw_time <- idw_time + idw_time_vec[k]


  # Run PPCA
  ppca_time_vec[k] <- system.time({
    ppca_imp <- PPCA(df_miss, nPcs = 20, transformation = sqrt,
                   reverse_transformation = function(x) x^2)
  })["elapsed"]
  ppca_time <- ppca_time + ppca_time_vec[k]


  # Run MissForest
  mf_time_vec[k] <- system.time({
    mf_imp <- MissForest(df_miss, ntree = 100, mtry = 100,
                     transformation = sqrt, reverse_transformation = function(x) x^2)
  })["elapsed"]
  mf_time <- mf_time + mf_time_vec[k]


  # Run MLR
  obs    <- df_miss[!is.na(df_miss$y), ]
  target <- df_miss[is.na(df_miss$y), ]
  mlr_time_vec[k] <- system.time({
    mlr_imp <- impute_MLR(obs, target, nn = 5, response = "y",
                    transformation = sqrt, reverse_transformation = function(x) x^2)
  })["elapsed"]
  mlr_time <- mlr_time + mlr_time_vec[k]


  # Fill in imputed values for this fold
  ence_pred[test_idx] <- ence_imp$y[test_idx]
  mice_pred[test_idx] <- mice_imp$y[test_idx]
  idw_pred[test_idx] <- idw_imp$y
  ppca_pred[test_idx] <- ppca_imp$y[test_idx]
  mf_pred[test_idx] <- mf_imp$y[test_idx]
  mlr_pred[test_idx] <- mlr_imp$y[test_idx]
}

true_y <- rain$y[observed_idx]
ence_y <- ence_pred[observed_idx]
mice_y <- mice_pred[observed_idx]
idw_y <- idw_pred[observed_idx]
ppca_y <- ppca_pred[observed_idx]
mf_y <- mf_pred[observed_idx]
mlr_y <- mlr_pred[observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(ence_y), !anyNA(mice_y), !anyNA(idw_y),!anyNA(ppca_y), !anyNA(mf_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed, na.rm = FALSE) {
  if (na.rm) {
    covered <- !is.na(pred)
    cat(method_name, "\n", sep = "")
    cat("Coverage:", sum(covered), "of", length(pred),
        sprintf("(%.1f%%)\n", 100 * mean(covered)))
    true <- true[covered]; pred <- pred[covered]
  } else {
    cat(method_name, "\n", sep = "")
  }
  rmse <- sqrt(mean((pred - true)^2))
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "Methods Comparison",
    cutoff      = cutoff,
    K           = K,
    seed        = 916,
    years       = years,
    n_stations  = length(keep),
    n_observed  = length(observed_idx),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all K folds
  ence_time = ence_time,
  mice_time = mice_time,
  idw_time = idw_time,
  ppca_time = ppca_time,
  mf_time = mf_time,
  mlr_time = mlr_time,
  ence_time_vec = ence_time_vec,
  mice_time_vec = mice_time_vec,
  idw_time_vec = idw_time_vec,
  ppca_time_vec = ppca_time_vec,
  mf_time_vec = mf_time_vec,
  mlr_time_vec = mlr_time_vec,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y = true_y,
  ence_y = ence_y,
  mice_y = mice_y,
  idw_y = idw_y,
  ppca_y = ppca_y,
  mf_y = mf_y,
  mlr_y = mlr_y,

  # Station Identifiers
  stno  = rain$stno[observed_idx],
  t     = rain$t[observed_idx],
  east  = rain$east[observed_idx],
  north = rain$north[observed_idx],
  fold  = fold_id
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, ence_y, "ENCE", ence_time)
  report_metrics(true_y, mice_y, "MICE", mice_time)
  report_metrics(true_y, idw_y, "IDW", idw_time)
  report_metrics(true_y, ppca_y, "PPCA", ppca_time)
  report_metrics(true_y, mf_y, "MissForest", mf_time)
  report_metrics(true_y, mlr_y, "MLR", mlr_time, na.rm = TRUE)

})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```