
# Filename: sklearn_test.Rmd
# Author: Marc Lane

# Methods Comparison Test

Run this .Rmd to reproduce the results from the first part of Section 4.3 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

# Use reticulate to call sklearn from R
python_path <- Sys.getenv("SKLEARN_PYTHON", unset = "/path/to/python3")
reticulate::use_python(python_path, required = TRUE)
reticulate::source_python("scripts/iterative_imputer.py")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")


# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_3_sklearn_%s.rds", tag)
log_file     <- sprintf("section_4_3_sklearn_%s.txt", tag)

```

# Imputation
 
Split data into 10 folds, mask observed entries and impute for each fold separately
```{r run, message=FALSE, results='hide'}
# Split the observed cells into K-folds
set.seed(916)
K <- 10
# Indexes of observed cells
observed_idx <- which(!is.na(rain$y))
# Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
fold_id      <- sample(rep(1:K, length.out = length(observed_idx)))
# Groups observed_idx into its fold
folds        <- split(observed_idx, fold_id)

# sklearn matrix setup
stno_levels <- sort(unique(rain$stno))
t_levels    <- sort(unique(rain$t))
row_idx_full <- match(rain$t, t_levels)
col_idx_full <- match(rain$stno, stno_levels)

# Initialising variables
sk_pred     <- rep(NA_real_, nrow(rain))
sk_time     <- 0
sk_time_vec <- numeric(K)

# Loop through each fold
for (k in seq_len(K)) {
  cat("Fold", k, "of", K, "\n")
  
  test_idx <- folds[[k]]
  
  # Hide the masked data for this fold
  df_miss <- rain
  df_miss$y[test_idx] <- NA

  # Prepare matrix for sklearn function
  X_matrix <- matrix(NA_real_, nrow = length(t_levels), ncol = length(stno_levels))
  row_idx  <- match(df_miss$t, t_levels)
  col_idx  <- match(df_miss$stno, stno_levels)
  X_matrix[cbind(row_idx, col_idx)] <- df_miss$y
  
  # run iterative imputer
  sk_time_vec[k] <- system.time({
    X_imputed <- run_iterative_imputer(X_matrix, n_nearest_features = 50L,
                                        max_iter = 10L, random_state = 916L)
   })["elapsed"]
  sk_time <- sk_time + sk_time_vec[k]

  cat("sklearn:", round(sk_time_vec[k], 1), "s   cumulative:", round(sk_time, 1), "s\n\n")

  sk_pred[test_idx] <- X_imputed[cbind(row_idx_full, col_idx_full)][test_idx]
}

true_y <- rain$y[observed_idx]
sk_y   <- sk_pred[observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(sk_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed, na.rm = FALSE) {
  if (na.rm) {
    covered <- !is.na(pred)
    cat(method_name, "\n", sep = "")
    cat("Coverage:", sum(covered), "of", length(pred),
        sprintf("(%.1f%%)\n", 100 * mean(covered)))
    true <- true[covered]; pred <- pred[covered]
  } else {
    cat(method_name, "\n", sep = "")
  }
  rmse <- sqrt(mean((pred - true)^2))
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "SKlearn only, Methods Comparison",
    cutoff      = cutoff,
    K           = K,
    seed        = 916,
    years       = years,
    n_stations  = length(keep),
    n_observed  = length(observed_idx),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),
  sk_time     = sk_time,
  sk_time_vec = sk_time_vec,
  true_y = true_y,
  sk_y   = sk_y,
  stno  = rain$stno[observed_idx],
  t     = rain$t[observed_idx],
  east  = rain$east[observed_idx],
  north = rain$north[observed_idx],
  fold  = fold_id
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, sk_y, "IterativeImputer", sk_time)

})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```