
# Filename: synchronous_vs_serial.Rmd
# Author: Marc Lane

# ENCE - Synchronous vs Serial Updates

Run this .Rmd to reproduce the results from Section 4.1 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/ENCE_serial.R")
source("scripts/ENCE_synchronous.R")

# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files <- sprintf("../met_eireann_daily_data/daily_rainfall_%d.csv", years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations   <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed) {
  rmse <- sqrt(mean((pred - true)^2))
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}

plot_obs_pred <- function(observed, predicted, method_name = "Method") {
  lims <- range(c(observed, predicted), na.rm = TRUE)
  plot(observed, predicted, xlim = lims, ylim = lims,
       xlab = "Observed", ylab = "Predicted",
       main = paste(method_name, ": Observed vs Predicted"),
       pch = 16, cex = 0.7)
  abline(0, 1, col = "red", lwd = 2)
  grid()
}
```

# Imputation
 
Split data into 10 folds, mask observed entries and impute for each fold separately
```{r run, message=FALSE, results='hide'}
# Split the observed cells into K-folds
set.seed(313)
K <- 10
# Indexes of observed cells
observed_idx <- which(!is.na(rain$y))
# Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
fold_id      <- sample(rep(1:K, length.out = length(observed_idx)))
# Groups observed_idx into its fold
folds        <- split(observed_idx, fold_id)

# Initialising variables
serial_pred <- rep(NA_real_, nrow(rain))
sync_pred   <- rep(NA_real_, nrow(rain))
serial_time <- 0
sync_time   <- 0
serial_rmse_list <- vector("list", K)
serial_val_list  <- vector("list", K)
sync_rmse_list <- vector("list", K)
sync_val_list  <- vector("list", K)
lambda_match <- 0L
lambda_valid <- 0L
alpha_match  <- 0L
alpha_valid  <- 0L

# Loop through each fold
for (k in seq_len(K)) {
  cat("Fold", k, "of", K, "\n")
  
  test_idx <- folds[[k]]
  
  # Hide the masked data for this fold
  df_miss <- rain
  df_miss$y[test_idx] <- NA
  
  # Run serial ENCE 
  serial_time <- serial_time + system.time({
    serial_imp <- ENCE_serial(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = test_idx,
                lambda_options = c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01),
                alpha_options  = c(0.1, 0.35, 0.65, 0.9)
                )
  })["elapsed"]

  # Run synchronous ENCE
  sync_time <- sync_time + system.time({
    sync_imp <- ENCE_sync(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = test_idx,
                     lambda_options = c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01),
                     alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                     n_workers = 16)
  })["elapsed"]

  # Fill in imputed values for this fold
  serial_pred[test_idx] <- serial_imp$y[test_idx]
  sync_pred[test_idx] <- sync_imp$y[test_idx]

  # Record RMSE lists for this fold
  serial_rmse_list[[k]] <- attr(serial_imp, "rmse_history")
  sync_rmse_list[[k]]   <- attr(sync_imp,   "rmse_history")
  serial_val_list[[k]]  <- attr(serial_imp, "val_rmse_history")
  sync_val_list[[k]]    <- attr(sync_imp,   "val_rmse_history")

  # Record hyperparameter values for this fold
  serial_l <- attr(serial_imp, "ls_history")[[2]]
  serial_a <- attr(serial_imp, "as_history")[[2]]
  sync_l <- attr(sync_imp,   "ls_history")[[2]]
  sync_a <- attr(sync_imp,   "as_history")[[2]]

  # Sum the number of matching hyperparameters and keep track
  vl <- !is.na(serial_l) & !is.na(sync_l)
  lambda_match <- lambda_match + sum(serial_l[vl] == sync_l[vl])
  lambda_valid <- lambda_valid + sum(vl)
  va <- !is.na(serial_a) & !is.na(sync_a)
  alpha_match  <- alpha_match  + sum(serial_a[va] == sync_a[va])
  alpha_valid  <- alpha_valid  + sum(va)
}

true_y   <- rain$y[observed_idx]
serial_y <- serial_pred[observed_idx]
sync_y   <- sync_pred[observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(serial_y), !anyNA(sync_y))

# Pad folds with shorter RMSE history with NA's, bind into a matrix 
pad <- function(lst) sapply(lst, function(v) c(v, rep(NA, max(lengths(lst)) - length(v))))
# Calculates the mean of the RMSE lists across folds
serial_rmse <- rowMeans(pad(serial_rmse_list), na.rm = TRUE)
sync_rmse   <- rowMeans(pad(sync_rmse_list),   na.rm = TRUE)
serial_val  <- rowMeans(pad(serial_val_list),  na.rm = TRUE)
sync_val    <- rowMeans(pad(sync_val_list),    na.rm = TRUE)

```

# Results

```{r serial_results, results='hold'}
# Print metrics
report_metrics(true_y, serial_y, "Serial", serial_time)
```

```{r sync_results, results='hold'}
# Print metrics
report_metrics(true_y, sync_y, "Synchronous", sync_time)
```

```{r scatterplots, results='hold'}
# Plot observed vs predicted scatterplots
plot_obs_pred(true_y, serial_y, "Serial")
plot_obs_pred(true_y, sync_y, "Synchronous")
```


```{r convergence_plot, results='hold'}
# Plot internal convergence for both methods
xlim_range <- c(1, max(length(serial_rmse), length(sync_rmse)))
plot(serial_rmse, type = "b", col = "blue", pch = 16,
     xlim = xlim_range,
     ylim = range(c(serial_rmse, sync_rmse)),
     xlab = "Cycle", ylab = "Internal RMSE (between cycles)",
     main = "Convergence: Serial vs Synchronous")
lines(sync_rmse, type = "b", col = "red", pch = 17)
legend("topright", legend = c("Sequential", "Synchronous"),
       col = c("blue", "red"), pch = c(16, 17), lty = 1)
```

```{r RMSE_plot, results='hold'}
# Plot validation RMSE for each method
xlim_range <- c(1, max(length(serial_val), length(sync_val)))
plot(serial_val, type = "b", col = "blue", pch = 16,
     xlim = xlim_range,
     ylim = range(c(serial_val, sync_val)),
     xlab = "Cycle", ylab = "Validation RMSE (masked cells)",
     main = "Validation RMSE: Sequential vs Synchronous")
lines(sync_val, type = "b", col = "red", pch = 17)
legend("topright", legend = c("Sequential", "Synchronous"),
       col = c("blue", "red"), pch = c(16, 17), lty = 1)
```

```{r hyperparameters, results='hold'}
# Print Hyperparameters
cat("Lambda agreement, pooled over", K, "folds\n")
cat("  ", lambda_match, "/", lambda_valid, "\n")
cat("  ", round(100 * lambda_match / lambda_valid, 1), "%\n")

cat("Alpha agreement, pooled over", K, "folds\n")
cat("  ", alpha_match, "/", alpha_valid, "\n")
cat("  ", round(100 * alpha_match / alpha_valid, 1), "%\n")
```
