# Filename: MICE_C_year_scaling_test.Rmd
# Author: Marc Lane

# MICE DURR - C Implementation Year Scaling Test

Run this .Rmd to reproduce the C implementation results for Section 4.2.6
of my thesis for the MICE algorithm: how does runtime scale with the number of
years of data?

NOTE: This is deliberately split out from MICE_sync_year_scaling_test.Rmd. Running
the OpenMP-parallel C implementation and then forking R parallel workers
(mclapply) in the same R session causes both to slow down - most likely because
the OpenMP thread pool from a prior C call is still alive when R forks the
workers. Submit this and the R-parallel version as two separate jobs/sessions
so neither implementation's timing is contaminated by the other.

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the
station completeness cutoff percentage. Each row is tagged with its calendar
year so we can build growing/sampled year subsets from a single load.

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/MICE_C.R")
source("scripts/MICE_elastic_net_C_implementation/elastic_net_R_glue.R")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years_full <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years_full)

# Read each year's file separately and tag rows with their calendar year,
# so later we can pull out arbitrary year subsets from a single loaded object
rain_list <- lapply(seq_along(files), function(i) {
  df <- read.csv(files[i])
  df$year <- years_full[i]
  df
})
rain <- do.call(rbind, rain_list)

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, year, y
rain <- rain[, c("stno", "east", "north", "t", "year", "rain")]
names(rain)[6] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station using the FULL 10 years. This means the set
# of stations kept (`keep`) is identical across every year-count in the sweep -
# only the time window changes, not the station panel.
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
names(obs_per_station) <- c("stno", "n_obs")
total_days <- length(unique(rain$t))
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep <- obs_per_station$stno[obs_per_station$completeness >= cutoff]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_2_6_MICE_C_year_scaling_%s.rds", tag)
log_file     <- sprintf("section_4_2_6_MICE_C_year_scaling_%s.txt", tag)
```

# Imputation

Sweep through year counts, and for each year count randomly sample that many
years (out of the 10 available) 3 times, masking 10% of observed cells each
rep - no K-fold CV, just enough reps to average out noise. Single imputation
(m = 1) since this is purely a timing sweep, not a beta/uncertainty analysis.

```{r run, message=FALSE, results='hide'}
year_values <- c(1, 2, 4, 6, 8, 10)
n_reps      <- 3
m           <- 1
p_fixed_stations <- keep

# Per-year-count mean timing, parallel to the old per-fold time_vec
C_time_mat <- matrix(NA_real_, length(year_values), n_reps)

# Pooled true/pred across every year count and rep, for report_metrics and scatterplots
C_true  <- c()
C_pred  <- c()
C_years <- c()  # number of years used, tagged per held-out cell, for plotting

# Record which calendar years were actually sampled each (year count, rep), for
# reproducibility / debugging - e.g. checking whether an unlucky sparse year
# distorted a particular rep's timing or accuracy
sampled_years_record <- vector("list", length(year_values))

# Record which stations get dropped for insufficient observations within a
# given sampled window (see guard below) - kept for diagnostics/reproducibility
dropped_stations_record <- vector("list", length(year_values))

for (ni in seq_along(year_values)) {
  Y <- year_values[ni]
  sampled_years_record[[ni]] <- vector("list", n_reps)
  dropped_stations_record[[ni]] <- vector("list", n_reps)

  for (r in seq_len(n_reps)) {
    set.seed(2000 + r)

    sampled_years <- sort(sample(years_full, Y))
    sampled_years_record[[ni]][[r]] <- sampled_years

    # Unmasked truth, captured before df_sub gets masked
    df_sub_truth <- rain[rain$stno %in% p_fixed_stations & rain$year %in% sampled_years, ]

    # Guard against stations with too few observations in THIS particular
    # sampled window: `keep`/p_fixed_stations was computed from 50% completeness
    # over the FULL 10 years, so a station can pass that cutoff overall while
    # having zero (or near-zero) observations in a specific 1-2 year sample -
    # e.g. a station installed partway through the record. A single all-NA (or
    # near all-NA) column poisons the correlation-based predictor selection for
    # every other column too. Tune min_obs_per_station against whatever overlap
    # the correlation step in MICE_impute.c actually needs to be numerically stable.
    min_obs_per_station <- 30
    obs_counts <- tapply(!is.na(df_sub_truth$y), df_sub_truth$stno, sum)
    sparse_stations <- as.numeric(names(obs_counts)[obs_counts < min_obs_per_station])
    if (length(sparse_stations)) {
      df_sub_truth <- df_sub_truth[!(df_sub_truth$stno %in% sparse_stations), ]
      cat("  dropped", length(sparse_stations), "station(s) with <", min_obs_per_station,
          "obs in this window:", paste(sparse_stations, collapse = ","), "\n")
    }
    dropped_stations_record[[ni]][[r]] <- sparse_stations

    df_sub <- df_sub_truth

    obs_idx  <- which(!is.na(df_sub$y))
    mask_idx <- sample(obs_idx, length(obs_idx) * 0.1)
    df_sub$y[mask_idx] <- NA

    true_y <- df_sub_truth$y[mask_idx]

    # Run C implementation MICE
    C_time_mat[ni, r] <- system.time({
      C_imp <- MICE_DURR_C(df_sub, response = "y", m = m,
                         max_impute_cycles = 5, hyp_cycles = 2,
                         init_method = "idw", max_EM_cycles = 8,
                         nclusters = 1, nthreads = 16, outfile = NULL,
                         lambda_options = lambda_options,
                         alpha_options  = alpha_options,
                         transformation = sqrt,
                         reverse_transformation = function(x) x^2)
    })["elapsed"]

    C_true  <- c(C_true, true_y)
    C_pred  <- c(C_pred, C_imp$y[mask_idx])
    C_years <- c(C_years, rep(Y, length(true_y)))

    cat("Years =", Y, "(", paste(sampled_years, collapse=","), ") rep", r, "done\n")
  }
}

# Per-year-count mean timing, parallel to the old per-fold *_time_vec
C_time_vec <- rowMeans(C_time_mat)

# Total time across the whole sweep
C_time <- sum(C_time_mat)

true_y <- C_true
C_y <- C_pred

stopifnot(!anyNA(C_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed) {
  rmse <- sqrt(mean((pred - true)^2))
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "MICE C implementation year scaling, Section 4.2.6",
    cutoff      = cutoff,
    years_full  = years_full,
    year_values = year_values,
    n_reps      = n_reps,
    m           = m,
    n_stations  = length(keep),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds
  C_time     = C_time,
  C_time_vec = C_time_vec,
  C_time_mat = C_time_mat,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y  = true_y,
  C_y     = C_y,
  C_years = C_years,

  # Which calendar years were sampled per (year count, rep)
  sampled_years_record = sampled_years_record,

  # Which stations were dropped for insufficient observations per (year count, rep)
  dropped_stations_record = dropped_stations_record
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, C_y, "C implementation", C_time)

  cat("Mean elapsed time by year count:\n")
  print(setNames(round(C_time_vec, 2), year_values))
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```
