# ENCE - C Implementation Test

In the directory `elastic_net_C_implementation`, I implemented my own version of glmnet (elastic net regressor) written in C with OpenMP parallelism. 

This file investigates if different versions of the C implementation with different optimizations
## Setup

Load the Met Eireann daily data (not available in this Github repo)

```{r setup, message=FALSE}
source("idw.R")
source("elastic_net_C_implementation/elastic_net_R_glue.R")
source("elastic_net_C_implementation_warm_start/elastic_net_R_glue.R")
source("elastic_net_C_implementation_active/elastic_net_R_glue.R")
source("elastic_net_C_implementation_optimize/elastic_net_R_glue.R")

# Load Met Eireann daily data and station geodata file
rain     <- read.csv("../met_eireann_daily_data/daily_rainfall_2016.csv")
stations <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)#
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Adds a completeness column = observed days / 366.
obs_per_station$completeness <- obs_per_station$n_obs / 366

# Apply the % cutoff
cutoff <- 0.50
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Then mask a fraction of the currently-observed cells
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```

## Imputation

Run the original glmnet, my C-version, the warm start optimization version, active set+warm start version and warmstart+activeset+ compiler flag and SIMD versions on the same masked data.


```{r run-glmnet, message=FALSE, results='hide'}
source("ENCE_glmnet.R")
time_glmnet <- system.time({
  imp_glmnet <- ENCE(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = masked_idx)
})
ext_rmse_glmnet <- attr(imp_glmnet, "ext_rmse_history")
```

```{r run-c_1core, message=FALSE, results='hide'}
source("ENCE_myelasticnet.R")
time_c_single <- system.time({
  imp_c_single <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx,
                nthreads = 1)
})
ext_rmse_c_single <- attr(imp_c_single, "ext_rmse_history")
```


```{r run-warmstart_1core, message=FALSE, results='hide'}
source("ENCE_myelasticnet_warmstart.R")
time_warmstart_single <- system.time({
  imp_warmstart_single <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx,
                nthreads = 1)
})
ext_rmse_warmstart_single <- attr(imp_warmstart_single, "ext_rmse_history")
```

```{r run-active_1core, message=FALSE, results='hide'}
source("ENCE_myelasticnet_active.R")
time_active_single <- system.time({
  imp_active_single <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx,
                nthreads = 1)
})
ext_rmse_active_single <- attr(imp_active_single, "ext_rmse_history")
```

```{r run-optimized_1core, message=FALSE, results='hide'}
source("ENCE_myelasticnet_optimize.R")
time_optimize_single <- system.time({
  imp_optimize_single <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx,
                nthreads = 1)
})
ext_rmse_optimize_single <- attr(imp_optimize_single, "ext_rmse_history")
```


## Results

### Timing

```{r timing}
cat("glmnet version:   ", round(time_glmnet["elapsed"], 2), "seconds\n")
cat("C version (1 thread):", round(time_c_single["elapsed"], 2), "seconds\n")
cat("Warmstart version (1 thread):", round(time_warmstart_single["elapsed"], 2), "seconds\n")
cat("Warmstart+active version:   ", round(time_active_single["elapsed"], 2), "seconds\n")
cat("Warmstart+active+optimized version:   ", round(time_optimize_single["elapsed"], 2), "seconds\n")
```

### Accuracy against truth

Final external RMSE (predicted vs. true rainfall at the masked locations):

```{r final-rmse}
cat("glmnet final ext_rmse: ", round(tail(ext_rmse_glmnet, 1), 4), "\n")
cat("C final ext_rmse (1 thread): ", round(tail(ext_rmse_c_single, 1), 4),"\n")
cat("Warm start final ext_rmse (1 thread): ", round(tail(ext_rmse_warmstart_single, 1), 4),"\n")
cat("Warm start+active final ext_rmse (1 thread): ", round(tail(ext_rmse_active_single, 1), 4),"\n")
cat("Warm start+active+optimized final ext_rmse (1 thread): ", round(tail(ext_rmse_optimize_single, 1), 4),"\n")
```

## Validation
```{r metrics}
y_true           <- rain$y[masked_idx]
glmnet_y         <- imp_glmnet$y[masked_idx]
ence_c1_y        <- imp_c_single$y[masked_idx]
ence_warm1_y     <- imp_warmstart_single$y[masked_idx]
ence_active1_y   <- imp_active_single$y[masked_idx]
ence_optimize1_y <- imp_optimize_single$y[masked_idx]

cat("glmnet\n")
cat("RMSE:",   sqrt(mean((glmnet_y - y_true)^2)), "\n")
cat("Bias %:", sum(glmnet_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((glmnet_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((glmnet_y - y_true)^2)) / mean(y_true), "\n\n")

cat("Naive C implementation\n")
cat("RMSE:",   sqrt(mean((ence_c1_y - y_true)^2)), "\n")
cat("Bias %:", sum(ence_c1_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((ence_c1_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((ence_c1_y - y_true)^2)) / mean(y_true), "\n\n")

cat("C (warm start)\n")
cat("RMSE:",   sqrt(mean((ence_warm1_y - y_true)^2)), "\n")
cat("Bias %:", sum(ence_warm1_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((ence_warm1_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((ence_warm1_y - y_true)^2)) / mean(y_true), "\n\n")

cat("C (active set)\n")
cat("RMSE:",   sqrt(mean((ence_active1_y - y_true)^2)), "\n")
cat("Bias %:", sum(ence_active1_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((ence_active1_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((ence_active1_y - y_true)^2)) / mean(y_true), "\n\n")

cat("C (optimized)\n")
cat("RMSE:",   sqrt(mean((ence_optimize1_y - y_true)^2)), "\n")
cat("Bias %:", sum(ence_optimize1_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((ence_optimize1_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((ence_optimize1_y - y_true)^2)) / mean(y_true), "\n\n")

cat("Warm vs cold C max abs diff:", max(abs(ence_warm1_y - ence_c1_y)), "\n")
cat("Active vs cold C max abs diff:", max(abs(ence_active1_y - ence_c1_y)), "\n")
cat("Optimized vs cold C max abs diff:", max(abs(ence_optimize1_y - ence_c1_y)), "\n")
cat("glmnet vs C max abs diff:", max(abs(glmnet_y - ence_c1_y)), "\n")
```