# ENCE - Sequential vs Synchronous Updates 
Imputing daily Met Eireann dataset, disregarding missing data and imputing on masked data to test RMSE

Investigating the convergence behaviour of the Elastic-net Chained Equations (ENCE) algorithm with Sequential vs Synchronous updates.

The original ENCE algorithm imputes each column of the data matrix $Z$ column-by-column within an outer cycle, where each station's regression model is fit using the most recently imputed values of all other stations.

The alternative algorithm considered here updates all stations within a cycle against a frozen snapshot of $Z$ taken at the start of that cycle, so the update order within a cycle no longer matters. The motivation for this change is to allow for parallelism.

Further details of ENCE can be found in the study at <https://doi.org/10.1002/joc.8513> The data that support the findings of this study are available upon reasonable request to the author(s) and with the permission of Met Éireann, Ireland's meteorological office.

## Setup

Load the Met Eireann daily data (not available in this Github repo) and source both ENCE variants.

```{r setup, message=FALSE}
source("idw.R")
source("ENCE_serial.R")     # sequential updates
source("ENCE_synchronous.R")   # synchronous updates

# Load Met Eireann daily data and station geodata file
rain     <- read.csv("../met_eireann_daily_data/daily_rainfall_2016.csv")
stations <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

head(rain)
str(rain)

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)#
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Adds a completeness column = observed days / 366.
obs_per_station$completeness <- obs_per_station$n_obs / 366

# Apply the % cutoff
cutoff <- 0.50
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Then mask a fraction of the currently-observed cells
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

truth   <- rain$y                     # ground truth at masked cells
df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```

## Imputation

Run both variants on the daily data:

```{r run-sequential_and_sychronous, message=FALSE, results='hide'}
source("ENCE_serial.R")
t0 <- Sys.time()
imp_seq <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = truth, masked_idx = masked_idx)
rmse_seq     <- attr(imp_seq, "rmse_history")
ext_rmse_seq <- attr(imp_seq, "ext_rmse_history")
cat(sprintf("Serial: %s\n", format(Sys.time() - t0)))

t0 <- Sys.time()
source("ENCE_synchronous.R")
imp_sync <- ENCE(df_miss, response = "y", max_cycles = 30,
                 init_method = "idw",
                 x_name = "east", y_name = "north", t_name = "t",
                 C = 15000,
                 transformation = sqrt,
                 reverse_transformation = function(x) x^2,
                 truth = truth, masked_idx = masked_idx)
rmse_sync <- attr(imp_sync, "rmse_history")
ext_rmse_sync <- attr(imp_sync, "ext_rmse_history")
cat(sprintf("Synchronous: %s\n", format(Sys.time() - t0)))
```

Plot 1: Internal Convergence

```{r plot-convergence, fig.width=8, fig.height=5}
xlim_range <- c(1, max(length(rmse_seq), length(rmse_sync)))

plot(rmse_seq, type = "b", col = "blue", pch = 16,
     xlim = xlim_range,
     ylim = range(c(rmse_seq, rmse_sync)),
     xlab = "Cycle", ylab = "Internal RMSE (between cycles)",
     main = "Internal convergence: sequential vs synchronous")
lines(rmse_sync, type = "b", col = "red", pch = 17)
legend("topright", legend = c("Sequential", "Synchronous"),
       col = c("blue", "red"), pch = c(16, 17), lty = 1)
```

Plot 2: RMSE Validation
```{r plot-RMSE, fig.width=8, fig.height=5}
xlim_range <- c(1, max(length(ext_rmse_seq), length(ext_rmse_sync)))

plot(ext_rmse_seq, type = "b", col = "blue", pch = 16,
     xlim = xlim_range,
     ylim = range(c(ext_rmse_seq, ext_rmse_sync)),
     xlab = "Cycle", ylab = "Validation RMSE (masked cells)",
     main = "Validation RMSE: sequential vs synchronous")
lines(ext_rmse_sync, type = "b", col = "red", pch = 17)
legend("topright", legend = c("Sequential", "Synchronous"),
       col = c("blue", "red"), pch = c(16, 17), lty = 1)
```

## Comparison
```{r comparison}
imputed_idx <- which(is.na(df_miss$y))
diffs <- imp_seq$y[imputed_idx] - imp_sync$y[imputed_idx]

cat(sprintf("Mean abs diff: %.4f mm\n", mean(abs(diffs))))
cat(sprintf("Max abs diff:  %.4f mm\n", max(abs(diffs))))
cat(sprintf("RMSE between methods: %.4f mm\n", sqrt(mean(diffs^2))))

```

## Agreement between sequential and synchronous imputations

For each cell that was missing in the input, plot the value the serial version filled in against the value the synchronous version filled in.

```{r serial-vs-sync}
plot(imp_seq$y[masked_idx], imp_sync$y[masked_idx],
     xlab = "Serial", ylab = "Synchronous")
abline(0, 1)
```

## Hyperparameter agreement

After cycle 2, both serial and synchronous methods freeze their hyperparameters for the remaining cycles. The percentages below give per-station agreement between the two methods on the lambda and alpha values they pick at cycle 2.

```{r hyperparameter-agreement}
l_seq  <- attr(imp_seq,  "ls_history")[[2]]
l_sync <- attr(imp_sync, "ls_history")[[2]]
a_seq  <- attr(imp_seq,  "as_history")[[2]]
a_sync <- attr(imp_sync, "as_history")[[2]]

valid <- !is.na(l_seq) & !is.na(l_sync)
cat(sprintf("Stations with matching lambda at cycle 2: %d / %d (%.1f%%) — %d NAs excluded\n",
            sum(l_seq[valid] == l_sync[valid]), sum(valid), 100 * mean(l_seq[valid] == l_sync[valid]), sum(!valid)))

valid <- !is.na(a_seq) & !is.na(a_sync)
cat(sprintf("Stations with matching alpha at cycle 2: %d / %d (%.1f%%) — %d NAs excluded\n",
            sum(a_seq[valid] == a_sync[valid]), sum(valid), 100 * mean(a_seq[valid] == a_sync[valid]), sum(!valid)))
```
