
# Filename: synchronous_vs_serial_mice.Rmd
# Author: Marc Lane

# ENCE - Synchronous vs Serial Updates

Run this .Rmd to reproduce the results from Section 4.1 of my thesis for the MICE algorithm

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/MICE_serial.R")
source("scripts/MICE_synchronous.R")

# Load Met Eireann daily data and station geodata file
years <- 2016
files <- sprintf("../met_eireann_daily_data/daily_rainfall_%d.csv", years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations   <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_1_MICE_sync_vs_serial_%s.rds", tag)
log_file     <- sprintf("section_4_1_MICE_sync_vs_serial_%s.txt", tag)
```

# Imputation
 
Split data into 10 folds, mask observed entries and impute for each fold separately
```{r run, message=FALSE, results='hide'}
# Split the observed cells into K-folds
set.seed(916)
K <- 10
# Indexes of observed cells
observed_idx <- which(!is.na(rain$y))
# Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
fold_id      <- sample(rep(1:K, length.out = length(observed_idx)))
# Groups observed_idx into its fold
folds        <- split(observed_idx, fold_id)

# Initialising variables
serial_pred <- rep(NA_real_, nrow(rain))
sync_pred   <- rep(NA_real_, nrow(rain))
serial_time <- 0
sync_time   <- 0
serial_rmse_list <- vector("list", K)
sync_rmse_list <- vector("list", K)

# Loop through each fold
for (k in seq_len(K)) {
  cat("Fold", k, "of", K, "\n")
  
  test_idx <- folds[[k]]
  
  # Hide the masked data for this fold
  df_miss <- rain
  df_miss$y[test_idx] <- NA
  
  # Run serial MICE
    serial_time <- serial_time + system.time({
    serial_imp <- MICE_DURR_serial(df_miss, response = "y", m = 10,
                       max_impute_cycles = 5, hyp_cycles = 2,
                       init_method = "idw", max_EM_cycles = 8,
                       nclusters = 10, outfile = "/dev/null",
                       lambda_options = c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01),
                       alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                       transformation = sqrt,
                       reverse_transformation = function(x) x^2)
  })["elapsed"]

  # Run synchronous MICE
  sync_time <- sync_time + system.time({
    sync_imp <- MICE_DURR_sync(df_miss, response = "y", m = 10,
                                 max_impute_cycles = 5, hyp_cycles = 2, 
                                 init_method = "idw", max_EM_cycles = 8, 
                                 nclusters = 10, outfile = "/dev/null",
                                 lambda_options = c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01),
                                 alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                                 transformation = sqrt,
                                 reverse_transformation = {function(x) x^2})
  })["elapsed"]

  # Fill in imputed values for this fold
  serial_pred[test_idx] <- serial_imp$y[test_idx]
  sync_pred[test_idx] <- sync_imp$y[test_idx]

  # Record RMSE lists for this fold
  serial_rmse_list[[k]] <- attr(serial_imp, "rmse_histories")
  sync_rmse_list[[k]]   <- attr(sync_imp,   "rmse_histories")

  cat("Cumulative  serial:", round(serial_time, 1), "s   sync:", round(sync_time, 1), "s\n")
}

true_y   <- rain$y[observed_idx]
serial_y <- serial_pred[observed_idx]
sync_y   <- sync_pred[observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(serial_y), !anyNA(sync_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed) {
  rmse <- sqrt(mean((pred - true)^2))
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "MICE synchronous vs serial, Section 4.1",
    cutoff      = cutoff,
    K           = K,
    seed        = 916,
    years       = years,
    n_stations  = length(keep),
    n_observed  = length(observed_idx),
    run_time    = Sys.time()
  ),

  # Timing in seconds summed over all K folds
  serial_time = serial_time,
  sync_time   = sync_time,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y   = true_y,
  serial_y = serial_y,
  sync_y   = sync_y,


  # Raw per-fold histories (length-K lists)
  serial_rmse_list = serial_rmse_list,
  sync_rmse_list   = sync_rmse_list,

  # Station Identifiers
  stno  = rain$stno[observed_idx],
  t     = rain$t[observed_idx],
  east  = rain$east[observed_idx],
  north = rain$north[observed_idx],
  fold  = fold_id
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, serial_y, "Serial", serial_time)
  report_metrics(true_y, sync_y, "Synchronous", sync_time)
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```

