
# Filename: ENCE_C_test.Rmd
# Author: Marc Lane

# ENCE - C implementation Test

Run this .Rmd to reproduce the results from the first part of Section 4.2 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/ENCE_serial.R")
source("scripts/ENCE_synchronous.R")
source("scripts/ENCE_C.R")
source("scripts/ENCE_elastic_net_C_implementation/elastic_net_R_glue.R")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years <- 2016
files <- sprintf("../met_eireann_daily_data/daily_rainfall_%d.csv", years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations   <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_2_ENCE_C_implementation_%s.rds", tag)
log_file     <- sprintf("section_4_2_ENCE_C_implementation_%s.txt", tag)
```

# Imputation
 
Split data into 10 folds, mask observed entries and impute for each fold separately
```{r run, message=FALSE, results='hide'}
# Split the observed cells into K-folds
set.seed(916)
K <- 10
# Indexes of observed cells
observed_idx <- which(!is.na(rain$y))
# Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
fold_id      <- sample(rep(1:K, length.out = length(observed_idx)))
# Groups observed_idx into its fold
folds        <- split(observed_idx, fold_id)

# Initialising variables
serial_pred <- rep(NA_real_, nrow(rain))
sync_pred   <- rep(NA_real_, nrow(rain))
C_pred   <- rep(NA_real_, nrow(rain))
serial_time <- 0
sync_time   <- 0
C_time   <- 0
serial_rmse_list <- vector("list", K)
serial_val_list  <- vector("list", K)
sync_rmse_list <- vector("list", K)
sync_val_list  <- vector("list", K)
C_rmse_list <- vector("list", K)
C_val_list  <- vector("list", K)
lambda_match <- 0L
lambda_valid <- 0L
alpha_match  <- 0L
alpha_valid  <- 0L
serial_lambda_final <- vector("list", K)
sync_lambda_final   <- vector("list", K)
C_lambda_final   <- vector("list", K)
serial_alpha_final  <- vector("list", K)
sync_alpha_final    <- vector("list", K)
C_alpha_final  <- vector("list", K)

# Loop through each fold
for (k in seq_len(K)) {
  cat("Fold", k, "of", K, "\n")
  
  test_idx <- folds[[k]]
  
  # Hide the masked data for this fold
  df_miss <- rain
  df_miss$y[test_idx] <- NA
  
  # Run serial ENCE 
  serial_time <- serial_time + system.time({
    serial_imp <- ENCE_serial(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = test_idx,
                lambda_options = lambda_options,
                alpha_options  = alpha_options
                )
  })["elapsed"]

  # Run synchronous ENCE
  sync_time <- sync_time + system.time({
    sync_imp <- ENCE_sync(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = test_idx,
                     lambda_options = lambda_options,
                     alpha_options  = alpha_options,
                     n_workers = 16)
  })["elapsed"]

  C_time <- C_time + system.time({
    C_imp <- ENCE_C(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = test_idx,
                lambda_options = lambda_options,
                alpha_options  = alpha_options,
                nthreads = 16)
  })["elapsed"]

  # Fill in imputed values for this fold
  serial_pred[test_idx] <- serial_imp$y[test_idx]
  sync_pred[test_idx] <- sync_imp$y[test_idx]
  C_pred[test_idx] <- C_imp$y[test_idx]

  # Record RMSE lists for this fold
  serial_rmse_list[[k]] <- attr(serial_imp, "rmse_history")
  sync_rmse_list[[k]]   <- attr(sync_imp,   "rmse_history")
  C_rmse_list[[k]]   <- attr(C_imp,   "rmse_history")
  serial_val_list[[k]]  <- attr(serial_imp, "val_rmse_history")
  sync_val_list[[k]]    <- attr(sync_imp,   "val_rmse_history")
  C_val_list[[k]]    <- attr(C_imp,   "val_rmse_history")

  # Record hyperparameter values for this fold
  serial_ls <- attr(serial_imp, "ls_history")
  serial_as <- attr(serial_imp, "as_history")
  sync_ls   <- attr(sync_imp,   "ls_history")
  sync_as   <- attr(sync_imp,   "as_history")
  C_ls   <- attr(C_imp,   "ls_history")
  C_as   <- attr(C_imp,   "as_history")
  serial_l <- serial_ls[[2]]
  serial_a <- serial_as[[2]]
  sync_l   <- sync_ls[[2]]
  sync_a   <- sync_as[[2]]
  C_l   <- C_ls[[2]]
  C_a   <- C_as[[2]]
  serial_lambda_final[[k]] <- serial_ls[[length(serial_ls)]]
  serial_alpha_final[[k]]  <- serial_as[[length(serial_as)]]
  sync_lambda_final[[k]]   <- sync_ls[[length(sync_ls)]]
  sync_alpha_final[[k]]    <- sync_as[[length(sync_as)]]
  C_lambda_final[[k]]   <- C_ls[[length(C_ls)]]
  C_alpha_final[[k]]    <- C_as[[length(C_as)]]

  # Sum the number of matching hyperparameters and keep track
  vl <- !is.na(serial_l) & !is.na(sync_l)
  lambda_match <- lambda_match + sum(serial_l[vl] == sync_l[vl])
  lambda_valid <- lambda_valid + sum(vl)
  va <- !is.na(serial_a) & !is.na(sync_a)
  alpha_match  <- alpha_match  + sum(serial_a[va] == sync_a[va])
  alpha_valid  <- alpha_valid  + sum(va)

  cat("Cumulative  R serial:", round(serial_time, 1), "s   R sync:", round(sync_time, 1), "s   C:", round(C_time, 1) ,"\n")
}

true_y   <- rain$y[observed_idx]
serial_y <- serial_pred[observed_idx]
sync_y   <- sync_pred[observed_idx]
C_y   <- C_pred[observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(serial_y), !anyNA(sync_y), !anyNA(C_y))

# Pad folds with shorter RMSE history with NA's, bind into a matrix 
pad <- function(lst) sapply(lst, function(v) c(v, rep(NA, max(lengths(lst)) - length(v))))
# Calculates the mean of the RMSE lists across folds
serial_rmse <- rowMeans(pad(serial_rmse_list), na.rm = TRUE)
sync_rmse   <- rowMeans(pad(sync_rmse_list),   na.rm = TRUE)
C_rmse   <- rowMeans(pad(C_rmse_list),   na.rm = TRUE)
serial_val  <- rowMeans(pad(serial_val_list),  na.rm = TRUE)
sync_val    <- rowMeans(pad(sync_val_list),    na.rm = TRUE)
C_val    <- rowMeans(pad(C_val_list),    na.rm = TRUE)

```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed) {
  rmse <- sqrt(mean((pred - true)^2))
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "ENCE Serial vs R parallel vs C, Section 4.2",
    cutoff      = cutoff,
    K           = K,
    seed        = 916,
    years       = years,
    n_stations  = length(keep),
    n_observed  = length(observed_idx),
    run_time    = Sys.time()
  ),

  # Timing in seconds summed over all K folds
  serial_time = serial_time,
  sync_time   = sync_time,
  C_time   = C_time,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y   = true_y,
  serial_y = serial_y,
  sync_y   = sync_y,
  C_y   = C_y,

  # Internal convergence RMSE (between cycles) averaged across folds per cycle
  serial_rmse = serial_rmse,
  sync_rmse   = sync_rmse,
  C_rmse   = C_rmse,

  # Validation RMSE on masked cells averaged across folds per cycle
  serial_val = serial_val,
  sync_val   = sync_val,
  C_val   = C_val,

  # Raw per-fold histories (length-K lists)
  serial_rmse_list = serial_rmse_list,
  sync_rmse_list   = sync_rmse_list,
  C_rmse_list   = C_rmse_list,
  serial_val_list  = serial_val_list,
  sync_val_list    = sync_val_list,
  C_val_list    = C_val_list,

  # Hyperparameter agreement, pooled over folds
  lambda_match = lambda_match,
  lambda_valid = lambda_valid,
  alpha_match  = alpha_match,
  alpha_valid  = alpha_valid,

  serial_lambda_final = serial_lambda_final,
  sync_lambda_final   = sync_lambda_final,
  C_lambda_final   = C_lambda_final,
  serial_alpha_final  = serial_alpha_final,
  sync_alpha_final    = sync_alpha_final,
  C_alpha_final    = C_alpha_final,

  # Station Identifiers
  stno  = rain$stno[observed_idx],
  t     = rain$t[observed_idx],
  east  = rain$east[observed_idx],
  north = rain$north[observed_idx],
  fold  = fold_id
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, serial_y, "Serial", serial_time)
  report_metrics(true_y, sync_y, "Synchronous", sync_time)
  report_metrics(true_y, C_y, "C implementation", C_time)

  cat("\nFinal lambda chosen (serial), pooled over", K, "folds\n")
  print(table(factor(unlist(serial_lambda_final), levels = sort(lambda_options))))
  cat("\nFinal lambda chosen (sync), pooled over", K, "folds\n")
  print(table(factor(unlist(sync_lambda_final), levels = sort(lambda_options))))
  cat("\nFinal lambda chosen (C), pooled over", K, "folds\n")
  print(table(factor(unlist(C_lambda_final), levels = sort(lambda_options))))
  cat("\nFinal alpha chosen (serial), pooled over", K, "folds\n")
  print(table(factor(unlist(serial_alpha_final), levels = sort(alpha_options))))
  cat("\nFinal alpha chosen (sync), pooled over", K, "folds\n")
  print(table(factor(unlist(sync_alpha_final), levels = sort(alpha_options))))
  cat("\nFinal alpha chosen (C), pooled over", K, "folds\n")
  print(table(factor(unlist(C_alpha_final), levels = sort(alpha_options))))

  cat("Lambda agreement, pooled over", K, "folds\n")
  cat("  ", lambda_match, "/", lambda_valid, "\n")
  cat("  ", round(100 * lambda_match / lambda_valid, 1), "%\n")
  cat("Alpha agreement, pooled over", K, "folds\n")
  cat("  ", alpha_match, "/", alpha_valid, "\n")
  cat("  ", round(100 * alpha_match / alpha_valid, 1), "%\n")
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```