
# Filename: MICE_C_test.Rmd
# Author: Marc Lane

# MICE DURR - C Implementation test

Run this .Rmd to reproduce the results from Section 4.2 of my thesis for the MICE algorithm

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/MICE_serial.R")
source("scripts/MICE_synchronous_parallel.R")
source("scripts/MICE_C.R")
source("scripts/MICE_elastic_net_C_implementation/elastic_net_R_glue.R")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years <- 2016
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_4_2_MICE_C_implementation_%s.rds", tag)
log_file     <- sprintf("section_4_2_MICE_C_implementation_%s.txt", tag)
```

# Imputation
 
Split data into 10 folds, mask observed entries and impute for each fold separately
```{r run, message=FALSE, results='hide'}
# Split the observed cells into K-folds
set.seed(916)
m <- 4
K <- 10
# Indexes of observed cells
observed_idx <- which(!is.na(rain$y))
# Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
fold_id      <- sample(rep(1:K, length.out = length(observed_idx)))
# Groups observed_idx into its fold
folds        <- split(observed_idx, fold_id)

# Initialising variables
serial_pred <- rep(NA_real_, nrow(rain))
sync_pred   <- rep(NA_real_, nrow(rain))
C_pred   <- rep(NA_real_, nrow(rain))
serial_time <- 0
sync_time   <- 0
C_time   <- 0
serial_rmse_list <- vector("list", K)
sync_rmse_list <- vector("list", K)
C_rmse_list <- vector("list", K)
serial_time_vec     <- numeric(K)
sync_time_vec <- numeric(K)
C_time_vec    <- numeric(K)
lambda_match <- 0L
lambda_valid <- 0L
alpha_match  <- 0L
alpha_valid  <- 0L
serial_lambda_final <- vector("list", K)
sync_lambda_final   <- vector("list", K)
C_lambda_final       <- vector("list", K)
serial_alpha_final  <- vector("list", K)
sync_alpha_final     <- vector("list", K)
C_alpha_final         <- vector("list", K)
C_beta_history_list  <- vector("list", K)

# Loop through each fold
for (k in seq_len(K)) {
  cat("Fold", k, "of", K, "\n")
  
  test_idx <- folds[[k]]
  
  # Hide the masked data for this fold
  df_miss <- rain
  df_miss$y[test_idx] <- NA
  
  # Run serial MICE
    serial_time_vec[k] <- system.time({
    serial_imp <- MICE_DURR_serial(df_miss, response = "y", m = m,
                       max_impute_cycles = 5, hyp_cycles = 2,
                       init_method = "idw", max_EM_cycles = 8,
                       nclusters = 1, outfile = "/dev/null",
                       lambda_options = lambda_options,
                       alpha_options  = alpha_options,
                       transformation = sqrt,
                       reverse_transformation = function(x) x^2)
  })["elapsed"]
  serial_time <- serial_time + serial_time_vec[k]

  # Run synchronous MICE
  sync_time_vec[k] <- system.time({
    sync_imp <- MICE_DURR_sync_par(df_miss, response = "y", m = m,
                                 max_impute_cycles = 5, hyp_cycles = 2, 
                                 init_method = "idw", max_EM_cycles = 8, 
                                 nclusters = 1, outfile = "/dev/null",
                                 n_workers = 16,
                                 lambda_options = lambda_options,
                                 alpha_options  = alpha_options,
                                 transformation = sqrt,
                                 reverse_transformation = {function(x) x^2})
  })["elapsed"]
  sync_time <- sync_time + sync_time_vec[k]

  C_time_vec[k] <- system.time({
    C_imp <- MICE_DURR_C(df_miss, response = "y", m = m,
                       max_impute_cycles = 5, hyp_cycles = 2,
                       init_method = "idw", max_EM_cycles = 8,
                       nclusters = 1, nthreads = 16, outfile = NULL,
                       lambda_options = lambda_options,
                       alpha_options  = alpha_options,
                       transformation = sqrt,
                       reverse_transformation = function(x) x^2)
  })["elapsed"]
  C_time <- C_time + C_time_vec[k]


  # Fill in imputed values for this fold
  serial_pred[test_idx] <- serial_imp$y[test_idx]
  sync_pred[test_idx] <- sync_imp$y[test_idx]
  C_pred[test_idx] <- C_imp$y[test_idx]

  # Record RMSE lists for this fold
  serial_rmse_list[[k]] <- attr(serial_imp, "rmse_histories")
  sync_rmse_list[[k]]   <- attr(sync_imp,   "rmse_histories")
  C_rmse_list[[k]]   <- attr(C_imp,   "rmse_histories")

  serial_lambdas_list <- attr(serial_imp, "lambdas_list")
  serial_alphas_list  <- attr(serial_imp, "alphas_list")
  sync_lambdas_list   <- attr(sync_imp,   "lambdas_list")
  sync_alphas_list    <- attr(sync_imp,   "alphas_list")
  C_lambdas_list      <- attr(C_imp,      "lambdas_list")
  C_alphas_list       <- attr(C_imp,      "alphas_list")

  serial_lambda_final[[k]] <- unlist(serial_lambdas_list)
  serial_alpha_final[[k]]  <- unlist(serial_alphas_list)
  sync_lambda_final[[k]]   <- unlist(sync_lambdas_list)
  sync_alpha_final[[k]]    <- unlist(sync_alphas_list)
  C_lambda_final[[k]]      <- unlist(C_lambdas_list)
  C_alpha_final[[k]]       <- unlist(C_alphas_list)

  C_beta_history_list[[k]] <- attr(C_imp, "beta_history_list")

  # Sum matching hyperparameters, per imputation copy (serial vs sync only)
  for (mi in seq_along(serial_lambdas_list)) {
    vl <- !is.na(serial_lambdas_list[[mi]]) & !is.na(sync_lambdas_list[[mi]])
    lambda_match <- lambda_match + sum(serial_lambdas_list[[mi]][vl] == sync_lambdas_list[[mi]][vl])
    lambda_valid <- lambda_valid + sum(vl)
    va <- !is.na(serial_alphas_list[[mi]]) & !is.na(sync_alphas_list[[mi]])
    alpha_match <- alpha_match + sum(serial_alphas_list[[mi]][va] == sync_alphas_list[[mi]][va])
    alpha_valid <- alpha_valid + sum(va)
  }

  cat("Cumulative  R serial:", round(serial_time, 1), "s   R sync:", round(sync_time, 1), "s   C:", round(C_time, 1) ,"\n")
}

true_y   <- rain$y[observed_idx]
serial_y <- serial_pred[observed_idx]
sync_y   <- sync_pred[observed_idx]
C_y   <- C_pred[observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(serial_y), !anyNA(sync_y), !anyNA(C_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed) {
  rmse <- sqrt(mean((pred - true)^2))
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "MICE C implementation test, Section 4.2",
    cutoff      = cutoff,
    K           = K,
    seed        = 916,
    years       = years,
    n_stations  = length(keep),
    n_observed  = length(observed_idx),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all K folds
  serial_time = serial_time,
  sync_time   = sync_time,
  C_time   = C_time,#
  serial_time_vec = serial_time_vec,
  sync_time_vec   = sync_time_vec,
  C_time_vec   = C_time_vec,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y   = true_y,
  serial_y = serial_y,
  sync_y   = sync_y,
  C_y   = C_y,


  # Raw per-fold histories (length-K lists)
  serial_rmse_list = serial_rmse_list,
  sync_rmse_list   = sync_rmse_list,
  C_rmse_list   = C_rmse_list,

  lambda_match = lambda_match,
  lambda_valid = lambda_valid,
  alpha_match  = alpha_match,
  alpha_valid  = alpha_valid,

  serial_lambda_final = serial_lambda_final,
  sync_lambda_final   = sync_lambda_final,
  C_lambda_final       = C_lambda_final,
  serial_alpha_final  = serial_alpha_final,
  sync_alpha_final     = sync_alpha_final,
  C_alpha_final         = C_alpha_final,

  C_beta_history_list = C_beta_history_list,

  # Station Identifiers
  stno  = rain$stno[observed_idx],
  t     = rain$t[observed_idx],
  east  = rain$east[observed_idx],
  north = rain$north[observed_idx],
  fold  = fold_id
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, serial_y, "Serial", serial_time)
  report_metrics(true_y, sync_y, "Synchronous", sync_time)
  report_metrics(true_y, C_y, "C", C_time)
  cat("\nLambda agreement, pooled over", K, "folds x", m, "imputations\n")
  cat("  ", lambda_match, "/", lambda_valid, "\n")
  cat("  ", round(100 * lambda_match / lambda_valid, 1), "%\n")
  cat("Alpha agreement, pooled over", K, "folds x", m, "imputations\n")
  cat("  ", alpha_match, "/", alpha_valid, "\n")
  cat("  ", round(100 * alpha_match / alpha_valid, 1), "%\n")
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```