# Filename: section_4_2_4_compiler_plots.Rmd
# Author: Marc Lane

# ENCE - Compilation Flags/Pragmas Sweep: Timing & Metrics

Reads the .rds / .txt pair produced by compiling_test.Rmd and reports
timing and final held-out metrics for the O0/O2/O3/march/unroll/flto/
noSIMD/production/fastmath builds. Convergence curves are intentionally
not reproduced here.

## Setup

```{r setup, message=FALSE}
# Must match the tag/results_file/log_file construction in compiling_test.Rmd
tag          <- sprintf("Compiler_flags_pragmas_Sweep")
results_file <- sprintf("section_4_2_compiler_sweep_%s.rds", tag)
log_file     <- sprintf("section_4_2_compiler_sweep_%s.txt", tag)

sched <- readRDS(results_file)

# Order matches the incremental build-up in compiling_test.Rmd:
# O0 -> O2 -> O3 -> march -> unroll -> flto -> production, with
# noSIMD and fastmath as ablations off the production config.
variants <- c("O0", "O2", "O3", "march", "unroll", "flto",
              "production", "noSIMD", "fastmath")
```

## Helper Functions

```{r helper_functions, message=FALSE}
compute_metrics <- function(true, pred) {
  rmse <- sqrt(mean((pred - true)^2))
  list(
    RMSE     = rmse,
    Bias_pct = sum(pred - true) / sum(true) * 100,
    R2       = 1 - sum((pred - true)^2) / sum((true - mean(true))^2),
    SRMSE    = rmse / mean(true)
  )
}

report_metrics <- function(true, pred, method_name, elapsed) {
  m <- compute_metrics(true, pred)
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", m$RMSE, "\n")
  cat("Bias %:", m$Bias_pct, "\n")
  cat("R2:    ", m$R2, "\n")
  cat("SRMSE: ", m$SRMSE, "\n\n")
}
```

## Metadata

```{r meta, results='hold'}
meta <- sched$meta
cat("ENCE Section 4.2.4 - Compiler Flags/Pragmas Sweep\n")
cat("cutoff:", meta$cutoff, " K:", meta$K, " seed:", meta$seed,
    " stations:", meta$n_stations, " observed:", meta$n_observed,
    " n_years:", length(meta$years),
    " years:", paste(meta$years, collapse = ", "), "\n")
cat("description:", meta$description, "\n")
cat("run_time:   ", format(meta$run_time), "\n")
if (!is.null(meta$hostname))       cat("hostname:   ", meta$hostname, "\n")
if (!is.null(meta$slurm_job_id))   cat("job:        ", meta$slurm_job_id, "\n")
if (!is.null(meta$slurm_nodelist)) cat("nodelist:   ", meta$slurm_nodelist, "\n")
if (!is.null(meta$r_version))      cat("R version:  ", meta$r_version, "\n")
cat("\n")
```

## Final Metrics

```{r metrics, results='hold'}
report_metrics(sched$true_y, sched$O0_y,         "O0",         sched$O0_time)
report_metrics(sched$true_y, sched$O2_y,         "O2",         sched$O2_time)
report_metrics(sched$true_y, sched$O3_y,         "O3",         sched$O3_time)
report_metrics(sched$true_y, sched$march_y,      "march",      sched$march_time)
report_metrics(sched$true_y, sched$unroll_y,     "unroll",     sched$unroll_time)
report_metrics(sched$true_y, sched$flto_y,       "flto",       sched$flto_time)
report_metrics(sched$true_y, sched$production_y, "production", sched$production_time)
report_metrics(sched$true_y, sched$noSIMD_y,     "noSIMD",     sched$noSIMD_time)
report_metrics(sched$true_y, sched$fastmath_y,   "fastmath",   sched$fastmath_time)

# Compact side-by-side comparison
preds <- setNames(lapply(variants, function(v) sched[[paste0(v, "_y")]]), variants)
summary_tbl <- data.frame(
  variant  = variants,
  time_s   = sapply(variants, function(v) sched[[paste0(v, "_time")]]),
  RMSE     = sapply(preds, function(p) compute_metrics(sched$true_y, p)$RMSE),
  Bias_pct = sapply(preds, function(p) compute_metrics(sched$true_y, p)$Bias_pct),
  R2       = sapply(preds, function(p) compute_metrics(sched$true_y, p)$R2),
  SRMSE    = sapply(preds, function(p) compute_metrics(sched$true_y, p)$SRMSE),
  row.names = NULL
)
print(summary_tbl, digits = 4)
```

## Timing

```{r timing, results='hold'}
cat("Per-fold wall time (s):\n")
time_vec_mat <- do.call(rbind, lapply(variants, function(v) sched[[paste0(v, "_time_vec")]]))
rownames(time_vec_mat) <- variants
print(round(time_vec_mat, 2))

cat("\nTotal time summed over", meta$K, "folds (s):\n")
totals <- sapply(variants, function(v) sched[[paste0(v, "_time")]])
print(round(totals, 1))

cat("\nPer-fold speedup vs O0 baseline, mean +/- sd (O0_time / variant_time):\n")
for (v in setdiff(variants, "O0")) {
  ratio <- sched$O0_time_vec / sched[[paste0(v, "_time_vec")]]
  cat(sprintf("O0 / %-10s: %6.3f +/- %.3f\n", v, mean(ratio), sd(ratio)))
}

cat("\nProduction-config ablations, per-fold ratio, mean +/- sd (variant_time / production_time):\n")
ratio_noSIMD   <- sched$noSIMD_time_vec   / sched$production_time_vec
ratio_fastmath <- sched$fastmath_time_vec / sched$production_time_vec
cat(sprintf("noSIMD   / production: %6.3f +/- %.3f  (>1 means SIMD pragmas were buying speed)\n",
            mean(ratio_noSIMD), sd(ratio_noSIMD)))
cat(sprintf("fastmath / production: %6.3f +/- %.3f  (<1 means -ffast-math bought further speed)\n",
            mean(ratio_fastmath), sd(ratio_fastmath)))
```

## Hyperparameters and full run log (from saved log file)

```{r log}
cat(readLines(log_file), sep = "\n")
```
