# Filename: section_4_2_5_schedule_plots.Rmd
# Author: Marc Lane

# ENCE - OpenMP Schedule Sweep: Timing & Metrics

Reads the .rds / .txt pair produced by schedule_test.Rmd and reports
timing and final held-out metrics for the dynamic/static/guided schedules.
Convergence curves are intentionally not reproduced here.

## Setup

```{r setup, message=FALSE}
# Must match the tag/results_file/log_file construction in schedule_test.Rmd
tag          <- sprintf("OpenMP_schedule_sweep")
results_file <- sprintf("section_4_2_schedule_sweep_%s.rds", tag)
log_file     <- sprintf("section_4_2_schedule_sweep_%s.txt", tag)

sched <- readRDS(results_file)
```

## Helper Functions

```{r helper_functions, message=FALSE}
compute_metrics <- function(true, pred) {
  rmse <- sqrt(mean((pred - true)^2))
  list(
    RMSE     = rmse,
    Bias_pct = sum(pred - true) / sum(true) * 100,
    R2       = 1 - sum((pred - true)^2) / sum((true - mean(true))^2),
    SRMSE    = rmse / mean(true)
  )
}

report_metrics <- function(true, pred, method_name, elapsed) {
  m <- compute_metrics(true, pred)
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", m$RMSE, "\n")
  cat("Bias %:", m$Bias_pct, "\n")
  cat("R2:    ", m$R2, "\n")
  cat("SRMSE: ", m$SRMSE, "\n\n")
}
```

## Metadata

```{r meta, results='hold'}
meta <- sched$meta
cat("ENCE Section 4.2.5 - OpenMP Schedule Sweep\n")
cat("cutoff:", meta$cutoff, " K:", meta$K, " seed:", meta$seed,
    " stations:", meta$n_stations, " observed:", meta$n_observed,
    " n_years:", length(meta$years),
    " years:", paste(meta$years, collapse = ", "), "\n")
cat("description:", meta$description, "\n")
cat("run_time:   ", format(meta$run_time), "\n")

# hostname/SLURM/R version aren't captured in this run's meta list
# (unlike the 4.2.1/4.2.2 runs) - print if present, skip otherwise
cat("hostname:   ", meta$hostname, "\n")
cat("job:        ", meta$slurm_job_id, "\n")
cat("nodelist:   ", meta$slurm_nodelist, "\n")
cat("R version:  ", meta$r_version, "\n")
cat("\n")
```

## Final Metrics

```{r metrics, results='hold'}
report_metrics(sched$true_y, sched$dynamic_y, "dynamic", sched$dynamic_time)
report_metrics(sched$true_y, sched$static_y,  "static",  sched$static_time)
report_metrics(sched$true_y, sched$guided_y,  "guided",  sched$guided_time)

# Compact side-by-side comparison
preds <- list(dynamic = sched$dynamic_y, static = sched$static_y, guided = sched$guided_y)
summary_tbl <- data.frame(
  schedule = names(preds),
  time_s   = c(sched$dynamic_time, sched$static_time, sched$guided_time),
  RMSE     = sapply(preds, function(p) compute_metrics(sched$true_y, p)$RMSE),
  Bias_pct = sapply(preds, function(p) compute_metrics(sched$true_y, p)$Bias_pct),
  R2       = sapply(preds, function(p) compute_metrics(sched$true_y, p)$R2),
  SRMSE    = sapply(preds, function(p) compute_metrics(sched$true_y, p)$SRMSE),
  row.names = NULL
)
print(summary_tbl, digits = 4)
```

## Timing

```{r timing, results='hold'}
cat("Per-fold wall time (s):\n")
print(rbind(dynamic = round(sched$dynamic_time_vec, 2),
            static  = round(sched$static_time_vec, 2),
            guided  = round(sched$guided_time_vec, 2)))

cat("\nTotal time summed over", meta$K, "folds (s):\n")
cat("dynamic:", round(sched$dynamic_time, 1),
    " static:", round(sched$static_time, 1),
    " guided:", round(sched$guided_time, 1), "\n")

ratio_dyn_static    <- sched$dynamic_time_vec / sched$static_time_vec
ratio_dyn_guided    <- sched$dynamic_time_vec / sched$guided_time_vec
ratio_static_guided <- sched$static_time_vec  / sched$guided_time_vec

cat("\nPer-fold time ratio, mean +/- sd (>1 means the first schedule was slower):\n")
cat("dynamic / static:", round(mean(ratio_dyn_static), 3),    "+/-", round(sd(ratio_dyn_static), 3), "\n")
cat("dynamic / guided:", round(mean(ratio_dyn_guided), 3),    "+/-", round(sd(ratio_dyn_guided), 3), "\n")
cat("static  / guided:", round(mean(ratio_static_guided), 3), "+/-", round(sd(ratio_static_guided), 3), "\n")
```

## Hyperparameters and full run log (from saved log file)

```{r log}
cat(readLines(log_file), sep = "\n")
```
