# Filename: section_4_3_timing_comparison.Rmd
# Author: Marc Lane

# Section 4.3 - Final Imputation Run: Computational Time Comparison

Reads the three result files produced by imputation_run.Rmd, imputation_run_mice.Rmd,
and imputation_run_sklearn.Rmd (full 10-year daily imputation, no masking, 3 reps each).

NOTE: R-parallel synchronous ENCE and MICE-DURR are still to be added to this run.
Once those .rds files exist, add two more rows to `summary_df` below (marked with a
comment) and re-knit.

## Setup

```{r setup, message=FALSE}
cutoff <- 0.50  # Must match the cutoff used in all three source Rmds

tag <- sprintf("cutoff%02d", round(cutoff * 100))

main_file    <- sprintf("section_4_3_Imputation_Run_%s.rds", tag)
mice_file    <- sprintf("section_4_3_Imputation_Run_mice_%s.rds", tag)
sklearn_file <- sprintf("section_4_3_Imputation_Run_sklearn_%s.rds", tag)

main_txt    <- sprintf("section_4_3_Imputation_Run_%s.txt", tag)
mice_txt    <- sprintf("section_4_3_Imputation_Run_mice_%s.txt", tag)
sklearn_txt <- sprintf("section_4_3_Imputation_Run_sklearn_%s.txt", tag)

main    <- readRDS(main_file)
mice    <- readRDS(mice_file)
sklearn <- readRDS(sklearn_file)
```

## Helper Functions

```{r helper_functions, message=FALSE}
report_timing <- function(method_name, elapsed_vec, n_missing, n_imputed_vec) {
  cat(method_name, "\n", sep = "")
  cat("Mean time:", round(mean(elapsed_vec), 2), "seconds  (sd:", round(sd(elapsed_vec), 2),
      "over", length(elapsed_vec), "reps)\n")
  cat("Per-rep times:", paste(round(elapsed_vec, 2), collapse = ", "), "\n")
  cat("Coverage:", round(mean(n_imputed_vec), 1), "of", n_missing,
      sprintf("(%.1f%%)\n\n", 100 * mean(n_imputed_vec) / n_missing))
}
```

## Run Metadata

```{r meta, results='hold'}
report_meta <- function(meta, label) {
  cat(label, "\n", sep = "")
  cat("cutoff:", meta$cutoff, " n_stations:", meta$n_stations,
      " n_missing:", meta$n_missing, " n_reps:", meta$n_reps,
      " years:", paste(range(meta$years), collapse = "-"), "\n")
  cat("description:", meta$description, "\n")
  cat("run_time:   ", format(meta$run_time), "\n")
  cat("hostname:   ", meta$hostname, " job:", meta$slurm_job_id,
      " nodelist:", meta$slurm_nodelist, "\n")
  cat("R version:  ", meta$r_version, "\n\n")
}

report_meta(main$meta, "Main run (ENCE / IDW / PPCA / MissForest / MLR)")
report_meta(mice$meta, "MICE-DURR run")
report_meta(sklearn$meta, "sklearn IterativeImputer run")

# Sanity check: all three runs must have loaded the same missingness pattern,
# otherwise the timings/coverage below are not directly comparable
n_missing <- main$meta$n_missing
stopifnot(
  mice$meta$n_missing    == n_missing,
  sklearn$meta$n_missing == n_missing
)
```

## Per-Method Timing Reports

```{r timing_reports, results='hold'}
report_timing("ENCE",       main$ence_time_vec, n_missing, main$ence_n_imputed)
report_timing("IDW",        main$idw_time_vec,  n_missing, main$idw_n_imputed)
report_timing("PPCA",       main$ppca_time_vec, n_missing, main$ppca_n_imputed)
report_timing("MissForest", main$mf_time_vec,   n_missing, main$mf_n_imputed)
report_timing("MLR",        main$mlr_time_vec,  n_missing, main$mlr_n_imputed)
report_timing("MICE-DURR",  mice$mice_time_vec, n_missing, mice$mice_n_imputed)
report_timing("sklearn IterativeImputer (HGBR)",
              sklearn$sklearn_time_vec, n_missing, sklearn$sklearn_n_imputed)
```

## Combined Summary Table

```{r summary_table}
summary_df <- data.frame(
  method = c("ENCE", "IDW", "PPCA", "MissForest", "MLR", "MICE-DURR",
             "sklearn (IterativeImputer/HGBR)"),
  mean_time_s = c(
    mean(main$ence_time_vec), mean(main$idw_time_vec), mean(main$ppca_time_vec),
    mean(main$mf_time_vec),   mean(main$mlr_time_vec),
    mean(mice$mice_time_vec), mean(sklearn$sklearn_time_vec)
  ),
  sd_time_s = c(
    sd(main$ence_time_vec), sd(main$idw_time_vec), sd(main$ppca_time_vec),
    sd(main$mf_time_vec),   sd(main$mlr_time_vec),
    sd(mice$mice_time_vec), sd(sklearn$sklearn_time_vec)
  ),
  coverage_pct = c(
    100 * mean(main$ence_n_imputed)       / n_missing,
    100 * mean(main$idw_n_imputed)        / n_missing,
    100 * mean(main$ppca_n_imputed)       / n_missing,
    100 * mean(main$mf_n_imputed)         / n_missing,
    100 * mean(main$mlr_n_imputed)        / n_missing,
    100 * mean(mice$mice_n_imputed)       / n_missing,
    100 * mean(sklearn$sklearn_n_imputed) / n_missing
  )
)

# --- ADD ROWS HERE once R-parallel synchronous ENCE / MICE-DURR results exist ---
# summary_df <- rbind(summary_df, data.frame(
#   method = c("ENCE (R sync)", "MICE-DURR (R sync)"),
#   mean_time_s = c(mean(ence_sync$...time_vec), mean(mice_sync$...time_vec)),
#   sd_time_s   = c(sd(ence_sync$...time_vec),   sd(mice_sync$...time_vec)),
#   coverage_pct = c(...)
# ))

summary_df <- summary_df[order(summary_df$mean_time_s), ]
summary_df$speedup_vs_slowest <- max(summary_df$mean_time_s) / summary_df$mean_time_s
rownames(summary_df) <- NULL

print(summary_df, row.names = FALSE)
```

## LaTeX Table (booktabs)

```{r latex_table, results='asis'}
library(knitr)
kable(summary_df, format = "latex", booktabs = TRUE, digits = 2,
      col.names = c("Method", "Mean time (s)", "SD (s)", "Coverage (\\%)",
                    "Speedup vs.\\ slowest"),
      caption = paste0("Section 4.3: mean wall-clock time over ",
                        main$meta$n_reps,
                        " reps for full 10-year daily imputation, ", tag, "."))
```

## Timing Bar Plot

```{r timing_barplot, fig.width=7, fig.height=5}
ord <- order(summary_df$mean_time_s)
bp <- barplot(summary_df$mean_time_s[ord], names.arg = summary_df$method[ord],
              las = 2, log = "y", ylab = "Mean time (s, log scale)",
              main = "Section 4.3: method wall-clock time comparison",
              col = "#4c72b0")
arrows(bp, pmax(summary_df$mean_time_s[ord] - summary_df$sd_time_s[ord], 1e-3),
       bp, summary_df$mean_time_s[ord] + summary_df$sd_time_s[ord],
       angle = 90, code = 3, length = 0.05)
```

## Raw Logs

```{r raw_logs, results='hold'}
cat(readLines(main_txt), sep = "\n")
cat("\n")
cat(readLines(mice_txt), sep = "\n")
cat("\n")
cat(readLines(sklearn_txt), sep = "\n")
```
