
# Filename: weak_scaling.Rmd
# Author: Marc Lane

# ENCE and MICE - C Implementation Weak Scaling

Run this .Rmd to reproduce the results from Section 4.2.2 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

source("scripts/idw.R")
source("scripts/ENCE_C.R")
source("scripts/ENCE_elastic_net_C_implementation/elastic_net_R_glue.R")
source("scripts/MICE_C.R")
source("scripts/MICE_elastic_net_C_implementation/elastic_net_R_glue.R")

# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Repeat over these number of threads
thread_grid <- c(1, 2, 4, 6, 8, 10, 12, 14, 16)
K <- 10

# Calculate the number of stations per thread for weak scaling
n_avail <- length(keep)
stations_per_thread <- floor(n_avail / max(thread_grid))
stopifnot(stations_per_thread >= 1)

set.seed(916)
station_order <- sample(keep)

# Setup output file names
tag          <- sprintf("C_weak")
results_file <- sprintf("section_4_2_ENCE_MICE_weak_scaling_%s.rds", tag)
log_file     <- sprintf("section_4_2_ENCE_MICE_weak_scaling_%s.txt", tag)
```

# Imputation

Split data into 10 folds, mask observed entries and impute for each fold separately
Repeat for different number of stations scaled proportionally with the number of threads (weak scaling)
```{r run, message=FALSE, results='hide'}

# Initialising variables (one entry per thread count)
ENCE_time <- setNames(numeric(length(thread_grid)), thread_grid)
MICE_time <- setNames(numeric(length(thread_grid)), thread_grid)
ENCE_pred <- rep(NA_real_, nrow(rain))
MICE_pred <- rep(NA_real_, nrow(rain))
ENCE_rmse_list <- vector("list", K)
ENCE_val_list  <- vector("list", K)
MICE_rmse_list <- vector("list", K)
ENCE_rmse_by_thread <- setNames(numeric(length(thread_grid)), thread_grid)
MICE_rmse_by_thread <- setNames(numeric(length(thread_grid)), thread_grid)
ENCE_time_mat <- matrix(NA_real_, nrow = K, ncol = length(thread_grid),
                         dimnames = list(NULL, as.character(thread_grid)))
MICE_time_mat <- matrix(NA_real_, nrow = K, ncol = length(thread_grid),
                         dimnames = list(NULL, as.character(thread_grid)))


for (n_threads in thread_grid) {
  cat("=== n_threads:", n_threads, " === \n")
  capture_run <- (n_threads == thread_grid[1])

  # Total number of stations for this number of threads
  n_stations_needed <- stations_per_thread * n_threads
  # Take a subset of stations
  stations_subset    <- station_order[1:n_stations_needed]
  # Get rainfall for this subset
  rain_sub           <- rain[rain$stno %in% stations_subset, ]

  # Indexes of observed cells
  observed_idx <- which(!is.na(rain_sub$y))
  set.seed(916)
  # Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
  fold_id <- sample(rep(1:K, length.out = length(observed_idx)))
  # Groups observed_idx into its fold
  folds   <- split(observed_idx, fold_id)

  ENCE_pred_thread <- rep(NA_real_, nrow(rain_sub))
  MICE_pred_thread <- rep(NA_real_, nrow(rain_sub))

  if (capture_run) {
    ENCE_pred <- rep(NA_real_, nrow(rain_sub))
    MICE_pred <- rep(NA_real_, nrow(rain_sub))
    capture_rain_sub    <- rain_sub
    capture_observed_idx <- observed_idx
    capture_fold_id      <- fold_id
  }

  # Loop through each fold
  for (k in seq_len(K)) {
    cat("Fold", k, "of", K, "\n")
  
    test_idx <- folds[[k]]
  
    # Hide the masked data for this fold
    df_miss <- rain_sub
    df_miss$y[test_idx] <- NA


    # Run synchronous ENCE
    ENCE_time_mat[k, as.character(n_threads)] <- system.time({
      ENCE_imp <- ENCE_C(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain_sub$y, masked_idx = test_idx,
                     lambda_options = c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01),
                     alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                     nthreads = n_threads)
    })["elapsed"]
    ENCE_time[as.character(n_threads)] <- ENCE_time[as.character(n_threads)] + ENCE_time_mat[k, as.character(n_threads)]


    # Run synchronous MICE
    MICE_time_mat[k, as.character(n_threads)] <- system.time({
      MICE_imp <- MICE_DURR_C(df_miss, response = "y", m = 1,
                                 max_impute_cycles = 5, hyp_cycles = 2, 
                                 init_method = "idw", max_EM_cycles = 8, 
                                 nclusters = 1,  nthreads = n_threads,
                                 outfile = "/dev/null",
                                 lambda_options = c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01),
                                 alpha_options  = c(0.1, 0.35, 0.65, 0.9),
                                 transformation = sqrt,
                                 reverse_transformation = {function(x) x^2})
    })["elapsed"]
    MICE_time[as.character(n_threads)] <- MICE_time[as.character(n_threads)] + MICE_time_mat[k, as.character(n_threads)]


    cat("Cumulative ENCE:", round(ENCE_time[as.character(n_threads)], 1), "s\n")
    cat("Cumulative MICE:", round(MICE_time[as.character(n_threads)], 1), "s\n")

    ENCE_pred_thread[test_idx] <- ENCE_imp$y[test_idx]
    MICE_pred_thread[test_idx] <- MICE_imp$y[test_idx]

    if (capture_run) {
      ENCE_pred[test_idx] <- ENCE_imp$y[test_idx]
      MICE_pred[test_idx] <- MICE_imp$y[test_idx]
      ENCE_rmse_list[[k]] <- attr(ENCE_imp, "rmse_history")
      ENCE_val_list[[k]]  <- attr(ENCE_imp, "val_rmse_history")
      MICE_rmse_list[[k]] <- attr(MICE_imp, "rmse_histories")
    }
  }

  ENCE_rmse_by_thread[as.character(n_threads)] <-
    sqrt(mean((ENCE_pred_thread[observed_idx] - rain_sub$y[observed_idx])^2))
  MICE_rmse_by_thread[as.character(n_threads)] <-
    sqrt(mean((MICE_pred_thread[observed_idx] - rain_sub$y[observed_idx])^2))
}


true_y <- capture_rain_sub$y[capture_observed_idx]
ENCE_y <- ENCE_pred[capture_observed_idx]
MICE_y <- MICE_pred[capture_observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(ENCE_y), !anyNA(MICE_y))

# Pad folds with shorter RMSE history with NA's, bind into a matrix 
pad <- function(lst) sapply(lst, function(v) c(v, rep(NA, max(lengths(lst)) - length(v))))
# Calculates the mean of the RMSE lists across folds
ENCE_rmse   <- rowMeans(pad(ENCE_rmse_list),   na.rm = TRUE)
ENCE_val    <- rowMeans(pad(ENCE_val_list),    na.rm = TRUE)

```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed) {
  rmse <- sqrt(mean((pred - true)^2))
  cat(method_name, "\n", sep = "")
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "ENCE/MICE C implementation Weak Scaling, Section 4.2.2",
    cutoff      = cutoff,
    K           = K,
    seed        = 916,
    years       = years,
    n_stations  = length(keep),
    n_observed  = length(capture_observed_idx),
    run_time    = Sys.time(),
    thread_grid = thread_grid,
    n_avail             = n_avail,
    stations_per_thread = stations_per_thread,
    station_order       = station_order,
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all K folds
  ENCE_time   = ENCE_time,
  MICE_time   = MICE_time,
  ENCE_time_mat = ENCE_time_mat,
  MICE_time_mat = MICE_time_mat,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y   = true_y,
  ENCE_y   = ENCE_y,
  MICE_y   = MICE_y,

  # Internal convergence RMSE (between cycles) averaged across folds per cycle
  ENCE_rmse   = ENCE_rmse,

  # Validation RMSE on masked cells averaged across folds per cycle
  ENCE_val   = ENCE_val,

  # Raw per-fold histories (length-K lists)
  ENCE_rmse_list   = ENCE_rmse_list,
  ENCE_val_list    = ENCE_val_list,
  MICE_rmse_list = MICE_rmse_list,

  ENCE_rmse_by_thread = ENCE_rmse_by_thread,
  MICE_rmse_by_thread = MICE_rmse_by_thread,

  # Station Identifiers
  stno  = capture_rain_sub$stno[capture_observed_idx],
  t     = capture_rain_sub$t[capture_observed_idx],
  east  = capture_rain_sub$east[capture_observed_idx],
  north = capture_rain_sub$north[capture_observed_idx],
  fold  = capture_fold_id
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, ENCE_y, "ENCE", ENCE_time["1"])
  report_metrics(true_y, MICE_y, "MICE", MICE_time["1"])
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```