# Filename: ENCE_N_scaling.Rmd
# Author: Marc Lane

# N scaling Methods Comparison Test

Run this .Rmd to reproduce the results from the first part of Section 4.3 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/ENCE_C_cov.R")
source("scripts/ENCE_elastic_net_C_implementation_cov/elastic_net_R_glue.R")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_cov_ENCE_N_scaling_%s.rds", tag)
log_file     <- sprintf("section_cov_ENCE_N_scaling_%s.txt", tag)
```

# Imputation

```{r run, message=FALSE, results='hide'}
N_values <- c(30, 90, 180, 365, 730, 1800, 3600)
n_reps   <- 3
p_fixed_stations <- keep

# Per-N mean timing (parallel to the old per-fold time_vec, now indexed by N_values)
ence_time_mat <- matrix(NA_real_, length(N_values), n_reps)

# Pooled true/pred across every N and rep, for report_metrics and scatterplots
ence_true <- c()
ence_pred <- c()
ence_N <- c()

for (ni in seq_along(N_values)) {
  N <- N_values[ni]

  for (r in seq_len(n_reps)) {
    set.seed(2000 + r)

    sampled_t <- sort(sample(unique(rain$t), N))

    # Unmasked truth, captured before df_sub gets masked
    df_sub_truth <- rain[rain$stno %in% p_fixed_stations & rain$t %in% sampled_t, ]
    df_sub <- df_sub_truth

    obs_idx  <- which(!is.na(df_sub$y))
    mask_idx <- sample(obs_idx, length(obs_idx) * 0.1)
    df_sub$y[mask_idx] <- NA

    true_y <- df_sub_truth$y[mask_idx]

    # Run ence
    ence_time_mat[ni, r] <- system.time({
      ence_imp <- ENCE_cov(df_sub, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = df_sub_truth$y, masked_idx = mask_idx,
                lambda_options = lambda_options,
                alpha_options  = alpha_options)
    })["elapsed"]
    ence_true <- c(ence_true, true_y)
    ence_pred <- c(ence_pred, ence_imp$y[mask_idx])
    ence_N <- c(ence_N, rep(N, length(true_y)))

    cat("N =", N, "rep", r, "done\n")
  }
}

# Per-N mean timing, parallel to the old per-fold *_time_vec
ence_time_vec <- rowMeans(ence_time_mat)

# Total time across the whole sweep
ence_time <- sum(ence_time_mat)

true_y <- ence_true   # same true values every method saw per rep
ence_y <- ence_pred

stopifnot(!anyNA(ence_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed, na.rm = FALSE) {
  if (na.rm) {
    covered <- !is.na(pred)
    cat(method_name, "\n", sep = "")
    cat("Coverage:", sum(covered), "of", length(pred),
        sprintf("(%.1f%%)\n", 100 * mean(covered)))
    true <- true[covered]; pred <- pred[covered]
  } else {
    cat(method_name, "\n", sep = "")
  }
  rmse <- sqrt(mean((pred - true)^2))
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "ENCE only, Methods Comparison N scaling, covariance",
    cutoff      = cutoff,
    years       = years,
    n_stations  = length(keep),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all K folds
  ence_time = ence_time,
  ence_time_vec = ence_time_vec,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y = true_y,
  ence_y = ence_y,

  ence_N = ence_N
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, ence_y, "ENCE", ence_time)
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```
# Filename: ENCE_N_scaling_cov.Rmd
# Author: Marc Lane

# N scaling Methods Comparison Test

Run this .Rmd to reproduce the results from the first part of Section 4.3 of my thesis

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/ENCE_C_cov.R")
source("scripts/ENCE_elastic_net_C_implementation_cov/elastic_net_R_glue.R")

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_cov_ENCE_N_scaling_%s.rds", tag)
log_file     <- sprintf("section_cov_ENCE_N_scaling_%s.txt", tag)
```

# Imputation

```{r run, message=FALSE, results='hide'}
N_values <- c(30, 90, 180, 365, 730, 1800, 3600)
n_reps   <- 3
p_fixed_stations <- keep

# Per-N mean timing (parallel to the old per-fold time_vec, now indexed by N_values)
ence_time_mat <- matrix(NA_real_, length(N_values), n_reps)

# Pooled true/pred across every N and rep, for report_metrics and scatterplots
ence_true <- c()
ence_pred <- c()
ence_N <- c()

for (ni in seq_along(N_values)) {
  N <- N_values[ni]

  for (r in seq_len(n_reps)) {
    set.seed(2000 + r)

    sampled_t <- sort(sample(unique(rain$t), N))

    # Unmasked truth, captured before df_sub gets masked
    df_sub_truth <- rain[rain$stno %in% p_fixed_stations & rain$t %in% sampled_t, ]
    df_sub <- df_sub_truth

    obs_idx  <- which(!is.na(df_sub$y))
    mask_idx <- sample(obs_idx, length(obs_idx) * 0.1)
    df_sub$y[mask_idx] <- NA

    true_y <- df_sub_truth$y[mask_idx]

    # Run ence
    ence_time_mat[ni, r] <- system.time({
      ence_imp <- ENCE_cov(df_sub, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = df_sub_truth$y, masked_idx = mask_idx,
                lambda_options = lambda_options,
                alpha_options  = alpha_options)
    })["elapsed"]
    ence_true <- c(ence_true, true_y)
    ence_pred <- c(ence_pred, ence_imp$y[mask_idx])
    ence_N <- c(ence_N, rep(N, length(true_y)))

    cat("N =", N, "rep", r, "done\n")
  }
}

# Per-N mean timing, parallel to the old per-fold *_time_vec
ence_time_vec <- rowMeans(ence_time_mat)

# Total time across the whole sweep
ence_time <- sum(ence_time_mat)

true_y <- ence_true   # same true values every method saw per rep
ence_y <- ence_pred

stopifnot(!anyNA(ence_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed, na.rm = FALSE) {
  if (na.rm) {
    covered <- !is.na(pred)
    cat(method_name, "\n", sep = "")
    cat("Coverage:", sum(covered), "of", length(pred),
        sprintf("(%.1f%%)\n", 100 * mean(covered)))
    true <- true[covered]; pred <- pred[covered]
  } else {
    cat(method_name, "\n", sep = "")
  }
  rmse <- sqrt(mean((pred - true)^2))
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  meta = list(
    description = "ENCE only, Methods Comparison N scaling, covariance",
    cutoff      = cutoff,
    years       = years,
    n_stations  = length(keep),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all K folds
  ence_time = ence_time,
  ence_time_vec = ence_time_vec,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y = true_y,
  ence_y = ence_y,

  ence_N = ence_N
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, ence_y, "ENCE", ence_time)
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```
