# Filename: missingness_mice.Rmd
# Author: Marc Lane

# MICE DURR Missingness Sweep - Covariance Update Accuracy (Section 4.3.1)

## Setup

Load the Met Eireann daily data (not available in this Github repo), apply the station completeness cutoff percentage

```{r setup, message=FALSE}

cutoff <- 0.50 # Set completeness cutoff percentage

source("scripts/idw.R")
source("scripts/MICE_C_cov.R")
source("scripts/MICE_elastic_net_C_implementation_cov/elastic_net_R_glue.R")

nodelist <- system('scontrol show hostnames "$SLURM_JOB_NODELIST"', intern = TRUE)

data_root <- Sys.getenv("MET_EIREANN_DATA", unset = "../../../met_eireann_daily_data")

lambda_options <- c(0.2, 0.15, 0.1, 0.05, 0.025, 0.01)
alpha_options  <- c(0.1, 0.35, 0.65, 0.9)

# Load Met Eireann daily data and station geodata file
years <- 2016:2025
files    <- sprintf("%s/daily_rainfall_%d.csv", data_root, years)
rain  <- do.call(rbind, lapply(files, read.csv))

stations <- read.csv(sprintf("%s/station_geodata.csv", data_root))

# Attach coordinates to each daily observation
missing_geo <- setdiff(unique(rain$stno), stations$stno)
if (length(missing_geo)) warning("No geodata for stno: ", paste(missing_geo, collapse=", "))
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Compute total number of days
total_days <- length(unique(rain$t))
# Adds a completeness column = observed days / total days.
obs_per_station$completeness <- obs_per_station$n_obs / total_days

# Apply the % cutoff
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Setup output file names
tag          <- sprintf("cutoff%02d", round(cutoff * 100))
results_file <- sprintf("section_cov_MICE_missingness_%s.rds", tag)
log_file     <- sprintf("section_cov_MICE_missingness_%s.txt", tag)
```

# Imputation

Split data into 10 folds, mask observed entries and impute for each fold separately.
Cluster is built once and reused across all K folds
```{r run, message=FALSE, results='hide'}
# Split the observed cells into K-folds
set.seed(916)
K <- 10
# Indexes of observed cells
observed_idx <- which(!is.na(rain$y))
# Shuffled repeating pattern 1,2,...,K as long as the number of observed cells
fold_id      <- sample(rep(1:K, length.out = length(observed_idx)))
# Groups observed_idx into its fold
folds        <- split(observed_idx, fold_id)

# Initialising variables
mice_pred <- rep(NA_real_, nrow(rain))
mice_time <- 0
mice_time_vec <- numeric(K)

# Pre-build cluster once, reused for every fold
library(parallel)
cl <- makeCluster(nodelist, outfile = sprintf("mice_cluster_%s.log", Sys.getenv("SLURM_JOB_ID")))
clusterExport(cl, c("rmse", "durr_impute_cycle_cov"))
so_path <- normalizePath("scripts/MICE_elastic_net_C_implementation_cov/MICE_impute_cov.so")
clusterExport(cl, "so_path")
clusterEvalQ(cl, dyn.load(so_path))
clusterEvalQ(cl, Sys.setenv(OPENBLAS_NUM_THREADS = "1", MKL_NUM_THREADS = "1"))

# Loop through each fold
for (k in seq_len(K)) {
  cat("Fold", k, "of", K, "\n")

  test_idx <- folds[[k]]

  # Hide the masked data for this fold
  df_miss <- rain
  df_miss$y[test_idx] <- NA

  # Run mice, covariance update version, m = 10 parallel across nodes
  mice_time_vec[k] <- system.time({
    mice_imp <- MICE_DURR_cov(df_miss, response = "y", m = 10,
                       max_impute_cycles = 5, hyp_cycles = 2,
                       init_method = "idw", max_EM_cycles = 8,
                       cl = cl, nthreads = 16, outfile = "",
                       lambda_options = lambda_options,
                       alpha_options  = alpha_options,
                       transformation = sqrt,
                       reverse_transformation = function(x) x^2)
  })["elapsed"]
  mice_time <- mice_time + mice_time_vec[k]
  cat("MICE:  ", round(mice_time_vec[k], 1), "s   cumulative:", round(mice_time, 1), "s\n")

  # Fill in imputed values for this fold
  mice_pred[test_idx] <- mice_imp$y[test_idx]
}

stopCluster(cl)

true_y <- rain$y[observed_idx]
mice_y <- mice_pred[observed_idx]

# Check every held-out cell got a prediction
stopifnot(!anyNA(mice_y))
```

## Helper Functions
```{r helper functions, message=FALSE}
report_metrics <- function(true, pred, method_name, elapsed, na.rm = FALSE) {
  if (na.rm) {
    covered <- !is.na(pred)
    cat(method_name, "\n", sep = "")
    cat("Coverage:", sum(covered), "of", length(pred),
        sprintf("(%.1f%%)\n", 100 * mean(covered)))
    true <- true[covered]; pred <- pred[covered]
  } else {
    cat(method_name, "\n", sep = "")
  }
  rmse <- sqrt(mean((pred - true)^2))
  cat("Time:  ", round(elapsed, 2), "seconds\n")
  cat("RMSE:  ", rmse, "\n")
  cat("Bias %:", sum(pred - true) / sum(true) * 100, "\n")
  cat("R2:    ", 1 - sum((pred - true)^2) / sum((true - mean(true))^2), "\n")
  cat("SRMSE: ", rmse / mean(true), "\n\n")
}
```

# Save Results

```{r save_results}
results <- list(
  # Run metadata, so this file identifies itself with no guessing
  meta = list(
    description = "MICE DURR covariance update, missingness sweep, Section 4.3.1",
    cutoff      = cutoff,
    K           = K,
    seed        = 916,
    years       = years,
    n_stations  = length(keep),
    n_observed  = length(observed_idx),
    run_time    = Sys.time(),
    hostname       = Sys.info()[["nodename"]],
    slurm_job_id   = Sys.getenv("SLURM_JOB_ID"),
    slurm_nodelist = Sys.getenv("SLURM_NODELIST"),
    r_version      = R.version.string
  ),

  # Timing in seconds summed over all K folds
  mice_time = mice_time,
  mice_time_vec = mice_time_vec,

  # Held-out predictions for observed-vs-predicted scatterplots
  true_y = true_y,
  mice_y = mice_y,

  # Station Identifiers
  stno  = rain$stno[observed_idx],
  t     = rain$t[observed_idx],
  east  = rain$east[observed_idx],
  north = rain$north[observed_idx],
  fold  = fold_id
)

saveRDS(results, results_file)
cat("Saved results to", results_file, "\n")
```

# Save printed log

```{r save_log}
log_lines <- capture.output({
  report_metrics(true_y, mice_y, "MICE", mice_time)
})

writeLines(log_lines, log_file)
cat(log_lines, sep = "\n")
cat("\nSaved log to", log_file, "\n")
```
