# ENCE - Daily Data test

I found previously that ENCE converges for the daily data, now investigated how well the imputation is performing.

Further details of ENCE can be found in the study at <https://doi.org/10.1002/joc.8513> The data that support the findings of this study are available upon reasonable request to the author(s) and with the permission of Met Éireann, Ireland's meteorological office.

## Setup

Load the Met Eireann daily data (not available in this Github repo)

```{r setup, message=FALSE}
source("idw.R")
source("ENCE_parallel.R")    # synchronous (parallel) updates

# Load Met Eireann daily data and station geodata file
rain     <- read.csv("../met_eireann_daily_data/daily_rainfall_2016.csv")
stations <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)#
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Adds a completeness column = observed days / 366.
obs_per_station$completeness <- obs_per_station$n_obs / 366

# Apply the % cutoff
cutoff <- 0.50
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Then mask a fraction of the currently-observed cells
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```

## Imputation
Comparing ENCE (with glmnet) with IDW and with ENCE with lag-1 temporal correlations in predictor matrix

```{r impute-idw, message=FALSE}
# IDW baseline
obs    <- df_miss[!is.na(df_miss$y), ]
target <- rain[masked_idx, ]
idw_out <- IDW_ST(obs, target, response = "y",
                  x_name = "east", y_name = "north", t_name = "t",
                  C = 15000,
                  transformation = sqrt,
                  reverse_transformation = function(x) x^2)
```

```{r impute-ENCE, message=FALSE}
# ENCE on the masked data
source("ENCE_parallel.R")  
imp <- ENCE(df_miss, response = "y", max_cycles = 30,
            init_method = "idw",
            x_name = "east", y_name = "north", t_name = "t",
            C = 15000,
            transformation = sqrt,
            reverse_transformation = function(x) x^2,
            truth = rain$y, masked_idx = masked_idx)
```

```{r impute-ENCE-temp, message=FALSE}
# ENCE on the masked data
source("ENCE_parallel_lag1.R")
imp_lag <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx)
```

## Validation
```{r metrics}
y_true <- rain$y[masked_idx]
ence_y <- imp$y[masked_idx]
lag_y  <- imp_lag$y[masked_idx]
idw_y  <- idw_out$y


cat("IDW\n")
cat("RMSE:", sqrt(mean((idw_y - y_true)^2)), "\n")
cat("Bias %:", sum(idw_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:", 1 - sum((idw_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n\n")

cat("ENCE\n")
cat("RMSE:", sqrt(mean((ence_y - y_true)^2)), "\n")
cat("Bias %:", sum(ence_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:", 1 - sum((ence_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n\n")

cat("ENCE + lag-1\n")
cat("RMSE:", sqrt(mean((lag_y - y_true)^2)), "\n")
cat("Bias %:", sum(lag_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:", 1 - sum((lag_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n\n")

cat("RMSE change from lag-1:",
    sqrt(mean((lag_y - y_true)^2)) - sqrt(mean((ence_y - y_true)^2)), "\n")
```