# ENCE - Overall Speedup

This file checks the speedup going from 1. The original ENCE implementation, 2. synchronous R-parallel updates (with glmnet) and 3. sychronous updates with my C implementation of glmnet

## Setup

Load the Met Eireann daily data (not available in this Github repo)

```{r setup, message=FALSE}
source("idw.R")
source("elastic_net_C_implementation_optimize/elastic_net_R_glue.R")

# Load Met Eireann daily data and station geodata file
rain     <- read.csv("../met_eireann_daily_data/daily_rainfall_2016.csv")
stations <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)#
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Adds a completeness column = observed days / 366.
obs_per_station$completeness <- obs_per_station$n_obs / 366

# Apply the % cutoff
cutoff <- 0.50
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Then mask a fraction of the currently-observed cells
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```

## Imputation
```{r run-original, message=FALSE, results='hide'}
source("ENCE_original.R")
time_original <- system.time({
  imp_original <- ENCE(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = masked_idx)
})
ext_rmse_original <- attr(imp_original, "ext_rmse_history")
```

```{r run-R, message=FALSE, results='hide'}
source("ENCE_R_parallel.R")
time_R <- system.time({
  imp_R <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx,
                n_cores = 6)
})
ext_rmse_R<- attr(imp_R, "ext_rmse_history")
```

```{r run-C, message=FALSE, results='hide'}
source("ENCE_C_parallel.R")
time_C <- system.time({
  imp_C <- ENCE(df_miss, response = "y", max_cycles = 30,
                init_method = "idw",
                x_name = "east", y_name = "north", t_name = "t",
                C = 15000,
                transformation = sqrt,
                reverse_transformation = function(x) x^2,
                truth = rain$y, masked_idx = masked_idx,
                nthreads = 6)
})
ext_rmse_C <- attr(imp_C, "ext_rmse_history")
```

## Results

### Timing

```{r timing}
cat("Original version:   ", round(time_original["elapsed"], 2), "seconds\n")
cat("R-Parallel version (6 threads):", round(time_R["elapsed"], 2), "seconds\n")
cat("C Implementation version (6 threads):", round(time_C["elapsed"], 2), "seconds\n")
```

```{r speedup}
t_orig <- time_original["elapsed"]
t_R    <- time_R["elapsed"]
t_C    <- time_C["elapsed"]

cat("Speedup, original -> R parallel:", round(t_orig / t_R, 1), "x\n")
cat("Speedup, original -> C:         ", round(t_orig / t_C, 1), "x\n")
cat("Speedup, R parallel -> C:       ", round(t_R / t_C, 1), "x\n")
```

### Accuracy against truth

Final external RMSE (predicted vs. true rainfall at the masked locations):

```{r final-rmse}
cat("R parallel final ext_rmse: ", round(tail(ext_rmse_R, 1), 4), "\n")
cat("C implementation final ext_rmse: ", round(tail(ext_rmse_C, 1), 4), "\n")
```

## Validation
```{r metrics}
y_true     <- rain$y[masked_idx]
original_y <- imp_original$y[masked_idx]
R_y        <- imp_R$y[masked_idx]
C_y        <- imp_C$y[masked_idx]

cat("Original (serial)\n")
cat("RMSE:",   sqrt(mean((original_y - y_true)^2)), "\n")
cat("Bias %:", sum(original_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((original_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((original_y - y_true)^2)) / mean(y_true), "\n\n")

cat("R parallel (synchronous, glmnet)\n")
cat("RMSE:",   sqrt(mean((R_y - y_true)^2)), "\n")
cat("Bias %:", sum(R_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((R_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((R_y - y_true)^2)) / mean(y_true), "\n\n")

cat("C implementation (synchronous)\n")
cat("RMSE:",   sqrt(mean((C_y - y_true)^2)), "\n")
cat("Bias %:", sum(C_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((C_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((C_y - y_true)^2)) / mean(y_true), "\n\n")

cat("R vs original max abs diff:", max(abs(R_y - original_y)), "\n")
cat("C vs original max abs diff:", max(abs(C_y - original_y)), "\n")
cat("C vs R max abs diff:",        max(abs(C_y - R_y)), "\n")
```