# ENCE - Method Comparisonls

This .Rmd does a comparison between different imputation methods

## Setup

Load the Met Eireann daily data (not available in this Github repo)

```{r setup, message=FALSE}
source("idw.R")
source("elastic_net_C_implementation/elastic_net_R_glue.R")

# Load Met Eireann daily data and station geodata file
rain     <- read.csv("../met_eireann_daily_data/daily_rainfall_2016.csv")
stations <- read.csv("../met_eireann_daily_data/station_geodata.csv")

# Attach coordinates to each daily observation
rain <- merge(rain, stations[, c("stno", "east", "north")], by = "stno")

# Reshape to the format the rest of the code expects: stno, east, north, t, y
rain <- rain[, c("stno", "east", "north", "t", "rain")]
names(rain)[5] <- "y"

# Sorts the table in place by stno, then t
rain <- rain[order(rain$stno, rain$t), ]

# Compute missingness per station
# Counts the number of observed days per station
obs_per_station <- aggregate(!is.na(rain$y), list(stno = rain$stno), sum)#
# Renames columns
names(obs_per_station) <- c("stno", "n_obs")
# Adds a completeness column = observed days / 366.
obs_per_station$completeness <- obs_per_station$n_obs / 366

# Apply the % cutoff
cutoff <- 0.50
keep   <- obs_per_station$stno[obs_per_station$completeness >= cutoff]
rain   <- rain[rain$stno %in% keep, ]

# Then mask a fraction of the currently-observed cells
set.seed(541)
observed_idx <- which(!is.na(rain$y))
masked_idx   <- sort(sample(observed_idx, length(observed_idx) * 0.25))

df_miss <- rain
df_miss$y[masked_idx] <- NA           # hide masked
```


## Run all Imputation Methods


```{r run-ENCE, message=FALSE, results='hide'}
source("ENCE.R")
time_ENCE <- system.time({
  imp_ENCE <- ENCE(df_miss, response = "y", max_cycles = 30,
                     init_method = "idw",
                     x_name = "east", y_name = "north", t_name = "t",
                     C = 15000,
                     transformation = sqrt,
                     reverse_transformation = function(x) x^2,
                     truth = rain$y, masked_idx = masked_idx,
                     nthreads = 6)
})
ext_rmse_ENCE <- attr(imp_ENCE, "ext_rmse_history")
```
```{r run-idw, message=FALSE}
source("idw.R")
obs    <- df_miss[!is.na(df_miss$y), ]
target <- rain[masked_idx, ]
time_idw <- system.time({
  imp_idw <- IDW_ST(obs, target, response = "y",
                  x_name = "east", y_name = "north", t_name = "t",
                  C = 15000,
                  transformation = sqrt,
                  reverse_transformation = function(x) x^2)
})
```
```{r run-MissForest, message=FALSE}
#source("MissForest.R")
#time_mf <- system.time({
#  imp_mf <- MissForest(df_miss, ntree = 100, mtry = 100,
#                     transformation = sqrt, reverse_transformation = function(x) x^2)
#})
```

```{r run-PPCA, message=FALSE}
source("PPCA.R")
time_ppca <- system.time({
  imp_ppca <- PPCA(df_miss, nPcs = 20, transformation = sqrt,
                   reverse_transformation = function(x) x^2)
})
```

```{r run-MLR, message=FALSE}
source("MLR.R")
obs    <- df_miss[!is.na(df_miss$y), ]
target <- df_miss[is.na(df_miss$y), ]
time_mlr <- system.time({
  imp_mlr <- impute_MLR(obs, target, nn = 5, response = "y")
})
```

```{r run-GAM, message=FALSE, results='hide'}
#source("GAM.R")
#time_gam <- system.time({
#  imp_gam <- ENCE(df_miss, response = "y", max_cycles = 30,
#                     init_method = "idw",
#                     x_name = "east", y_name = "north", t_name = "t",
#                     C = 15000,
#                     transformation = sqrt,
#                     reverse_transformation = function(x) x^2,
#                     truth = rain$y, masked_idx = masked_idx,
#                     nthreads = 6)
#})

## Results
```{r metrics}
y_true     <- rain$y[masked_idx]
ENCE_y <- imp_ENCE$y[masked_idx]
idw_y  <- imp_idw$y
#mf_y <- imp_mf$y[masked_idx]
ppca_y <- imp_ppca$y[masked_idx]
mlr_y <- imp_mlr$y[masked_idx]
#gam_y <- imp_gam$y[masked_idx]
```


```{r ENCE-results, results='hold'}
cat("ENCE:\n")
cat("Time: ", round(time_ENCE["elapsed"], 2), "seconds\n")
#cat("C implementation final ext_rmse: ", round(tail(ext_rmse_ENCE, 1), 4), "\n")
cat("RMSE:",   sqrt(mean((ENCE_y - y_true)^2)), "\n")
cat("Bias %:", sum(ENCE_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((ENCE_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((ENCE_y - y_true)^2)) / mean(y_true), "\n\n")
```

```{r idw-results, results='hold'}
cat("IDW:\n")
cat("Time", round(time_idw["elapsed"], 2), "seconds\n")
cat("RMSE:",   sqrt(mean((idw_y - y_true)^2)), "\n")
cat("Bias %:", sum(idw_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((idw_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((idw_y - y_true)^2)) / mean(y_true), "\n\n")
```

```{r MissForest-results, results='hold'}
#cat("MissForest:\n")
#cat("Time", round(time_mf["elapsed"], 2), "seconds\n")
#cat("RMSE:",   sqrt(mean((mf_y - y_true)^2)), "\n")
#cat("Bias %:", sum(mf_y - y_true) / sum(y_true) * 100, "\n")
#cat("R2:",     1 - sum((mf_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
#cat("SRMSE:",  sqrt(mean((mf_y - y_true)^2)) / mean(y_true), "\n\n")
```

```{r PPCA-results, results='hold'}
cat("PPCA:\n")
cat("Time", round(time_ppca["elapsed"], 2), "seconds\n")
cat("RMSE:",   sqrt(mean((ppca_y - y_true)^2)), "\n")
cat("Bias %:", sum(ppca_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((ppca_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((ppca_y - y_true)^2)) / mean(y_true), "\n\n")
```

```{r MLR-results, results='hold'}
cat("MLR:\n")
cat("Time", round(time_mlr["elapsed"], 2), "seconds\n")
cat("RMSE:",   sqrt(mean((mlr_y - y_true)^2)), "\n")
cat("Bias %:", sum(mlr_y - y_true) / sum(y_true) * 100, "\n")
cat("R2:",     1 - sum((mlr_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
cat("SRMSE:",  sqrt(mean((mlr_y - y_true)^2)) / mean(y_true), "\n\n")
```

```{r GAM-results, results='hold'}
#cat("GAM:\n")
#cat("Time", round(time_gam["elapsed"], 2), "seconds\n")
#cat("RMSE:",   sqrt(mean((gam_y - y_true)^2)), "\n")
#cat("Bias %:", sum(gam_y - y_true) / sum(y_true) * 100, "\n")
#cat("R2:",     1 - sum((gam_y - y_true)^2) / sum((y_true - mean(y_true))^2), "\n")
#cat("SRMSE:",  sqrt(mean((gam_y - y_true)^2)) / mean(y_true), "\n\n")
```